---
title: "Appendix — ADMB → RTMB Code "
subtitle: "Tracking translation details while bridging ADMB to RTMB"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-tools: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    documentclass: scrartcl
execute:
  echo: false
  warning: false
  message: false
editor: source
editor_options: 
  chunk_output_type: console
---

# Purpose

This standalone appendix tracks items to verify or refactor as the model is bridged from **ADMB** to **RTMB**. Each entry records the ADMB behavior, intended RTMB behavior, priority, and links/tests.

# Model implementation

The RTMB model code is implemented in the `R/` directory, with the main entry point 
in `R/Rpm.R`. 


## Comparison of ADMB and RTMB by components

The initial steps to bridge ADMB to RTMB involve identifying key components and their behaviors. The following table summarizes the components, their ADMB behavior, intended RTMB behavior, and current status.

```{r}
#| echo: false
#| results: asis
#| message: false
#| warning: false
#| error: false
rm(list = ls())
# source(here::here("R/Rpm_annotated.R"))
source(here::here("R/Rpm.R"))
# eb_bts/ pm$eb_bts

gt_compare_table(rtmb, pm, tolerance = 1e-2, sort_by_diff=FALSE)
```

## Tuning


## Issues register for next steps

```{r}
#| label: issues-register
# Self-contained chunk (defines has_gt here to avoid scoping issues)
has_gt <- requireNamespace("gt", quietly = TRUE)

needs <- c("dplyr","tibble","stringr","glue")
ok <- vapply(needs, function(p) requireNamespace(p, quietly = TRUE), logical(1))
if (!all(ok)) stop("Please install packages: ", paste(needs[!ok], collapse = ", "))
invisible(lapply(c("dplyr","tibble"), library, character.only = TRUE))

issues_path <- file.path("inst","issues.csv")
if (file.exists(issues_path)) {
  issues <- read.csv(issues_path, stringsAsFactors = FALSE)
} else {
  issues <- tibble::tibble(
    id        = c("ISS-001","ISS-002","ISS-003"),
    component = c("Selectivity (temporal devs)",
                  "Selectivity (age-smoothing)",
                  "Survey likelihood (lognormal scaling)"),
    admb_behavior = c(
      "First-difference penalty in **time** with year 1 anchored to 0 (Δ log s_{y,a} = log s_{y,a} - log s_{y-1,a}, with log s_{0,a}≡0).",
      "First-difference penalty in **age** with minimum age anchored to 0 (Δ log s_{y,a} = log s_{y,a} - log s_{y,a-1}, with log s_{y,a_min-1}≡0).",
      "ADMB often uses −(n/2)log(2πσ^2) − Σ (log(O_i)−log(E_i))^2 / (2σ^2). Verify whether σ is on log or natural scale and whether bias-corrected ln residuals are used."
    ),
    rtmb_behavior = c(
      "Replicate identical time FD penalty in R/RTMB; ensure first row uses 0 as the lag.",
      "Replicate identical age FD penalty with explicit 0 anchor at a_min in each year.",
      "Match exact scaling, including any constants and phase-dependent weights; confirm whether to include the normalizing constant in objective function."
    ),
    status    = c("Pending","Pending","Pending"),
    priority  = c("High","Medium","High"),
    test_ref  = c("test-fd-penalty-time","test-fd-penalty-age","test-survey-lognorm")
  )
}

if (has_gt) {
  dom_status <- function(x) dplyr::case_when(
    x %in% c("Pending","Blocked") ~ 0.9,
    x %in% c("In progress")       ~ 0.5,
    TRUE                          ~ 0.1
  )
  dom_pri <- function(x) dplyr::case_when(
    x == "High"   ~ 0.9,
    x == "Medium" ~ 0.5,
    TRUE          ~ 0.2
  )
  plt <- gt::gt(issues)
  plt <- gt::data_color(
    plt, columns = c(status),
    fn = function(vals) grDevices::colorRampPalette(c("white","gold","orangered"))(100)[
      ceiling(pmax(pmin(dom_status(vals),1),0)*99)+1
    ],
    autocolor_text = TRUE
  )
  plt <- gt::data_color(
    plt, columns = c(priority),
    fn = function(vals) grDevices::colorRampPalette(c("white","lightblue","royalblue"))(100)[
      ceiling(pmax(pmin(dom_pri(vals),1),0)*99)+1
    ],
    autocolor_text = TRUE
  )
  gt::tab_caption(plt, "ADMB → RTMB Issues Register")
} else {
  knitr::kable(issues, caption = "ADMB → RTMB Issues Register")
}
```

### Auto-generated anchors

```{r}
#| label: issues-anchors
#| results: asis
stopifnot(requireNamespace("glue", quietly = TRUE))
issues |>
  dplyr::rowwise() |>
  dplyr::mutate(block = {
    glue::glue("
#### {id}: {component}

**ADMB behavior**: {admb_behavior}

**RTMB target**: {rtmb_behavior}

**Status**: {status} &nbsp; | &nbsp; **Priority**: {priority}

See test/ref: `#{test_ref}`.
")
  }) |>
  dplyr::pull(block) |>
  cat(sep = "\n\n")
```


::: callout-tip
**Usage**  
Edit the CSV at `inst/issues.csv`. Re‑render to update the register and anchors.
:::



```{r}
needs <- c("dplyr","tibble","stringr","glue")
ok <- vapply(needs, function(p) requireNamespace(p, quietly = TRUE), logical(1))
if (!all(ok)) stop("Please install packages: ", paste(needs[!ok], collapse = ", "))
invisible(lapply(c("dplyr","tibble"), library, character.only = TRUE))

has_gt <- requireNamespace("gt", quietly = TRUE)
```

## Model changes

### First‑difference selectivity penalty (temporal) {#test-fd-penalty-time}

We need the penalty used in ADMB where the **first year is compared to zero** (not to a previous estimated year). For a matrix of log‑selectivities \( \log s_{y,a} \) (years \(y=1,\ldots,T\), ages \(a=1,\ldots,A\)), the penalty is:

$$
\mathcal{P}_{\text{FD,time}} \,=\, w \sum_{a=1}^{A}
\Big[
\big(\log s_{1,a}-0\big)^2
+ \sum_{y=2}^{T}\big(\log s_{y,a}-\log s_{y-1,a}\big)^2
\Big].
$$

#### RTMB/R implementation

```{r}
#| echo: true
# log_sel: matrix [T x A] of log-selectivities (year rows, age cols)
# w: scalar or length-A vector of weights
fd_penalty_time <- function(log_sel, w = 1) {
  stopifnot(is.matrix(log_sel))
  T <- nrow(log_sel); A <- ncol(log_sel)
  if (length(w) == 1) w <- rep(w, A)
  stopifnot(length(w) == A)

  diffs <- rbind(log_sel[1, , drop = FALSE] - 0,   # year 1 vs 0
                 apply(log_sel, 2, diff))          # y>=2 diffs
  sum(colSums((sqrt(w) * diffs)^2))
}
```

### Sanity check

```{r}
#| echo: true
set.seed(42)
T <- 5; A <- 3
L <- matrix(rnorm(T*A, sd=0.2), T, A); w <- c(1, 2, 0.5)

fd_loop <- function(L, w){
  acc <- 0
  for (a in seq_len(ncol(L))) {
    acc <- acc + w[a]*(L[1,a]-0)^2
    for (y in 2:nrow(L)) acc <- acc + w[a]*(L[y,a]-L[y-1,a])^2
  }
  acc
}

all.equal(fd_penalty_time(L, w), fd_loop(L, w))
```

### First‑difference selectivity penalty (age) {#test-fd-penalty-age}

For within-year smoothing over age with a **zero anchor at the minimum age**:

$$
\mathcal{P}_{\text{FD,age}} \,=\, w \sum_{y=1}^{T}
\Big[
\big(\log s_{y,a_{\min}}-0\big)^2
+ \sum_{a=a_{\min}+1}^{A}\big(\log s_{y,a}-\log s_{y,a-1}\big)^2
\Big].
$$

#### RTMB/R implementation

```{r}
#| echo: true
# log_sel: matrix [T x A] of log-selectivities (year rows, age cols)
# a_min: column index of minimum age within 'log_sel' (default = 1)
fd_penalty_age <- function(log_sel, w = 1, a_min = 1) {
  stopifnot(is.matrix(log_sel))
  T <- nrow(log_sel); A <- ncol(log_sel)
  stopifnot(a_min >= 1, a_min <= A)
  if (length(w) == 1) w <- rep(w, T)
  stopifnot(length(w) == T)

  # Build diffs across age with anchor at a_min (assumes a_min=1; if not, set zeros below a_min)
  diffs <- matrix(0, nrow = T, ncol = A)
  diffs[, a_min] <- log_sel[, a_min] - 0
  if (a_min < A) {
    for (a in (a_min+1):A) {
      diffs[, a] <- log_sel[, a] - log_sel[, a-1]
    }
  }
  sum(rowSums((sqrt(w) * diffs)^2))
}
```

## Reference: survey lognormal scaling (consistency check) {#test-survey-lognorm}

When porting lognormal survey likelihoods, confirm whether ADMB’s objective includes the normalizing constant and whether \(\sigma\) is on the **log** scale:

$$
\ell(\sigma) = -\sum_i \frac{(\log O_i - \log E_i)^2}{2\sigma^2}
\; -\; n\log\sigma \; -\; \frac{n}{2}\log(2\pi).
$$

```{r}
#| echo: true
lognorm_nll <- function(obs, exp, sigma_log, include_const = TRUE) {
  r <- log(pmax(obs, .Machine$double.eps)) - log(pmax(exp, .Machine$double.eps))
  n <- length(r)
  nll <- sum(r^2) / (2 * sigma_log^2)
  if (include_const) nll <- nll + n*log(sigma_log) + n*0.5*log(2*pi)
  nll
}
```
