This section documents two distinct age-composition product issues considered after the 2024
assessment: the method used to assign ages to individual otoliths and the
spatio-temporal method used to expand BTS age samples into annual proportions
at age. The first concerns traditional microscope ages (TMA) versus ages
predicted by Fourier-transform near-infrared spectroscopy (FT-NIRS). The
second concerns replacement of the prior BTS age-composition estimator with a
tinyVAST implementation. In the second case, the survey observations remain
the same; the estimated annual age-composition data product derived from those
observations has changed because the expansion method was revised. These
changes occur at different stages of the data workflow and should be evaluated
separately.

### FTNIRS age-data {#sec-ftnirs-age-data}

The Alaska Fisheries Science Center's Age and Growth Program has produced more
than 460,000 walleye pollock age estimates since 1971. From 2005 through 2024,
annual production averaged 11,141 pollock ages and ranged from 8,408 to 14,441.
Across Alaska pollock stocks, inter-reader precision was relatively high: the
overall average coefficient of variation was 4.28%, percent agreement was
68.6%, and reader and tester estimates generally differed by one year or less
year. Eighty percent of pollock aged by the program were 3--10 years old, and
the maximum observed age was 31 years. Radiometric analysis supports the
accuracy of the program's ageing criteria through age 8, while marginal
increment analyses, year-class tracking, hard-structure comparisons, and
captive-rearing studies provide additional corroboration. Older ages occur
more frequently in the Bering Sea and Aleutian Islands than in the Gulf of
Alaska, and Bering Sea pollock generally grow more slowly but attain greater
lengths and ages. Female pollock also become larger at age than males beginning
at approximately age 5 [@matta2026agegrowth, pp. 33--42]. These production and
precision statistics summarize AFSC pollock ageing across Alaska stocks rather
than the eastern Bering Sea alone; the eastern Bering Sea assessment is the
most data-rich pollock assessment.

These results describe traditional microscope ageing (TMA), which assigns age
by preparing an otolith and visually counting annual growth zones. FT-NIRS is
being developed as a rapid secondary ageing method that depends on reference
ages produced through traditional microscopic examination
[@matta2026agegrowth]. FT-NIRS measures the wavelengths and amount of
near-infrared light absorbed by an otolith and predicts age from its spectral
and chemical signal using a calibration model. It can process otoliths
substantially faster, but its predicted ages are methodologically different
from direct microscope readings ([NOAA Fisheries age-method
overview](https://www.fisheries.noaa.gov/feature-story/age-and-growth-homework-determining-how-old-fish-are)). The established scale, precision, and
corroboration of the TMA series therefore provide important context for
evaluating whether FT-NIRS predictions preserve the age structure and cohort
signals used by the assessment.

The 2025 CIE discussion paper compared the two methods using pollock data. For
the same design-based BTS abundance-at-age inputs, TMA estimates showed a high
degree of cohort consistency, whereas FT-NIRS estimates showed lower cohort
consistency; the assessment authors described the consequence of that
difference as unresolved. The comparison also identified differences between
TMA- and FT-NIRS-based proportions at age for the fishery, BTS, and ATS. Model
tests therefore treated FT-NIRS ages as a separate data source and considered
both global and fleet-specific FT-NIRS age-error matrices. See the official
[FT-NIRS data consistency
comparisons](https://noaa-afsc.github.io/EBS_pollock/doc/CIE.html#ft-nirs-data-consistency-comparisons)
for the cohort plots, composition comparisons, data coverage, and alternative
age-error treatments.

The FT-NIRS issue concerns age assignment and associated ageing error. The BTS
bridge below concerns the subsequent expansion of age observations across
space and time. Its scope is the expansion method; a separate analysis
compares TMA with FT-NIRS.

### Impact of revised BTS age-composition estimates {#sec-bts-age-data-bridge}

This section isolates the effect of replacing the original bottom-trawl
survey (BTS) estimates of proportions at age in `BTSProp.csv` with the revised
estimates in `tinyVAST_props_new.csv`. The age-composition data product changes
while the underlying survey data remain the same. Both ADMB runs use the
complete 2025 input set and the
same model specification, executable, starting parameter file, and tinyVAST
2025 BTS age-composition estimate. The sensitivity restores the original
1982-2024 BTS proportion estimates while holding the 2025 tinyVAST row
constant. The comparison therefore addresses the historical estimator change.

The 2024 assessment used the previous spatio-temporal expansion of BTS age
samples. The revised series uses an improved implementation in `tinyVAST`, a
multivariate spatio-temporal generalized linear mixed-model framework that can
model age groups jointly, share information across locations and years, and
area-expand predictions to annual abundance and proportions at age
[@thorson2019spatiotemporalcomposition; @thorson2025tinyvast]. The
[tinyVAST age-composition expansion
example](https://vast-lib.github.io/tinyVAST/articles/web_only/age_composition_expansion.html)
documents this workflow for EBS pollock age data. This methodological update
changes the estimated BTS proportions at age supplied to the assessment; the
controlled runs below quantify the resulting assessment-model effect while
holding the other inputs and model configuration fixed.

```{r}
#| label: bts-age-data-issues-load
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(gt)
  library(tibble)
  library(tidyr)
})

theme_set(ggthemes::theme_few())

report_source <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
age_data_dir <- file.path(
  dirname(report_source), "data", "EBSpollock_Sept_2026"
)

read_report_csv <- function(filename) {
  path <- file.path(age_data_dir, filename)
  if (!file.exists(path)) {
    stop("Missing September 2026 report input: ", path)
  }
  as_tibble(read.csv(path, check.names = FALSE, stringsAsFactors = FALSE))
}

drop_all_na_cols <- function(data) {
  data |>
    select(where(~!all(is.na(.x))))
}

gt_report <- function(data, ...) {
  data <- drop_all_na_cols(data)
  table <- gt(data, ...)

  integer_cols <- names(data)[tolower(names(data)) %in% c("year", "age")]
  if (length(integer_cols) > 0) {
    table <- table |>
      fmt_number(columns = all_of(integer_cols), decimals = 0, use_seps = FALSE)
  }

  table
}

bts_age_input_summary <- read_report_csv("input_summary.csv")
bts_age_input_comparison <- read_report_csv("age_data_comparison.csv")
bts_age_diagnostics <- read_report_csv("diagnostics.csv")
bts_age_comparison <- read_report_csv("comparison.csv")

bts_age_input_extremes <- bts_age_input_comparison |>
  summarise(
    maximum_increase = max(percent_change),
    maximum_decrease = min(percent_change),
    maximum_percentage_point_change = max(
      abs(100 * (tinyvast_proportion - original_proportion))
    )
  )

bts_age_result_summary <- bts_age_comparison |>
  group_by(quantity) |>
  summarise(
    terminal_year = max(year),
    terminal_percent_difference = percent_difference[which.max(year)],
    maximum_absolute_percent_difference = max(abs(percent_difference)),
    year_of_maximum_difference = year[which.max(abs(percent_difference))],
    .groups = "drop"
  ) |>
  mutate(
    quantity = recode(
      quantity,
      spawning_biomass = "Spawning biomass",
      recruitment_age1 = "Age-1 recruitment",
      predicted_catch = "Predicted catch"
    )
  )
```

The controlled input comparison changes the historical BTS age-composition
product while holding the 2025 age composition and every other assessment
input constant (@tbl-bts-age-data-contrast).

```{r}
#| label: tbl-bts-age-data-contrast
#| tbl-cap: "Controlled input contrast for the revised BTS age-composition product. The 2025 BTS age-composition estimate and all other assessment inputs are identical between runs."
bts_age_input_summary |>
  gt_report() |>
  tab_header(title = "BTS Age-Data Contrast")
```

The derived-product comparison uses the `both` region from `BTSProp.csv` as
the original estimated series and `tinyVAST_props_new.csv` as the revised
estimated series
(@fig-bts-age-input-change). Across the 42 matched survey years, relative
changes range from
`r sprintf("%.1f%%", bts_age_input_extremes$maximum_decrease)` to
`r sprintf("%+.1f%%", bts_age_input_extremes$maximum_increase)`; the largest
absolute change is
`r sprintf("%.2f percentage points", bts_age_input_extremes$maximum_percentage_point_change)`.
The calculation uses the original estimated proportion as the denominator and
describes the revision to the derived age-composition product rather than a
change in the survey observations or its influence on model results. The
revised 2025 row is excluded because the original file ends in 2024.

```{r}
#| label: fig-bts-age-input-change
#| fig-cap: "Percent change in the estimated BTS proportions at age from the original BTSProp.csv data product to the revised tinyVAST_props_new.csv data product for each matched historical survey year. The underlying survey observations are unchanged. Positive bars indicate a larger tinyVAST estimate and negative bars indicate a smaller tinyVAST estimate. Age 15 is the plus group; the BTS survey was absent in 2020."
#| fig-alt: "Small-multiple bar chart comparing two age-composition data products derived from the same BTS observations for 42 matched survey years from 1982 through 2024, excluding 2020. Each panel shows relative percent change from the original estimated proportions to the tinyVAST estimates for ages 1 through 15. Bars above zero indicate increases and bars below zero indicate decreases. Changes range from about minus 12 percent to plus 55 percent, with the largest increase at age 12 in 2023."
#| fig-width: 13
#| fig-height: 10
bts_age_input_comparison |>
  ggplot(aes(age, percent_change, fill = direction)) +
  geom_hline(yintercept = 0, color = "grey35", linewidth = 0.3) +
  geom_col(width = 0.82) +
  facet_wrap(vars(year), ncol = 7) +
  scale_x_continuous(breaks = c(1, 5, 10, 15)) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  scale_fill_manual(
    values = c("Decrease" = "#2166AC", "Increase" = "#B2182B")
  ) +
  labs(
    x = "Age (15 is plus group)",
    y = "Change from prior proportion at age",
    fill = "Direction"
  ) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

Both runs reached maximum absolute gradients below 0.001. The likelihood
components in @tbl-bts-age-data-diagnostics document the fits and support a
structured contrast between scenarios with different observed BTS age values.

```{r}
#| label: tbl-bts-age-data-diagnostics
#| tbl-cap: "ADMB fit diagnostics for the matched 2025 model with tinyVAST BTS proportions and with original BTSProp proportions for 1982-2024. Negative-log-likelihood values diagnose each data version; cross-data model selection requires a common observation set and likelihood definition."
bts_age_diagnostics |>
  gt_report() |>
  fmt_number(
    columns = c(total_nll, bts_index_nll, bts_age_nll),
    decimals = 3
  ) |>
  fmt_scientific(columns = maximum_gradient, decimals = 2) |>
  cols_label(
    scenario = "Data version",
    total_nll = "Total NLL",
    bts_index_nll = "BTS index NLL",
    bts_age_nll = "BTS age NLL",
    maximum_gradient = "Maximum |gradient|"
  ) |>
  tab_header(title = "BTS Age-Data Bridge Diagnostics")
```

The revision changes terminal spawning biomass by
`r sprintf("%+.3f%%", bts_age_result_summary$terminal_percent_difference[bts_age_result_summary$quantity == "Spawning biomass"])`.
The largest spawning-biomass difference is
`r sprintf("%.3f%%", bts_age_result_summary$maximum_absolute_percent_difference[bts_age_result_summary$quantity == "Spawning biomass"])`
in `r bts_age_result_summary$year_of_maximum_difference[bts_age_result_summary$quantity == "Spawning biomass"]`.
Terminal age-1 recruitment changes by
`r sprintf("%+.3f%%", bts_age_result_summary$terminal_percent_difference[bts_age_result_summary$quantity == "Age-1 recruitment"])`,
which is the largest recruitment difference in the series. Terminal predicted
catch changes by
`r sprintf("%+.3f%%", bts_age_result_summary$terminal_percent_difference[bts_age_result_summary$quantity == "Predicted catch"])`.
The terminal and maximum differences for each assessment quantity are
summarized in @tbl-bts-age-data-results.

```{r}
#| label: tbl-bts-age-data-results
#| tbl-cap: "Relative effects of replacing the original historical BTS proportions with tinyVAST proportions. Percent differences are tinyVAST minus original, divided by the absolute original estimate."
bts_age_result_summary |>
  rename(
    result_quantity = quantity,
    result_terminal_year = terminal_year
  ) |>
  gt_report() |>
  fmt_number(
    columns = c(
      terminal_percent_difference,
      maximum_absolute_percent_difference
    ),
    decimals = 3
  ) |>
  cols_label(
    result_quantity = "Quantity",
    result_terminal_year = "Terminal year",
    terminal_percent_difference = "Terminal difference (%)",
    maximum_absolute_percent_difference = "Maximum absolute difference (%)",
    year_of_maximum_difference = "Year of maximum"
  ) |>
  tab_header(title = "Effects of tinyVAST Historical BTS Proportions")
```

Across years, spawning-biomass and predicted-catch differences remain small,
while age-1 recruitment shows the larger year-to-year response
(@fig-bts-age-data-bridge).

```{r}
#| label: fig-bts-age-data-bridge
#| fig-cap: "Percent differences in spawning biomass, age-1 recruitment, and predicted catch between the 2025 model using tinyVAST BTS proportions and the sensitivity using original BTSProp proportions for 1982-2024. The horizontal line marks zero difference."
#| fig-alt: "Three-panel line graph of percent differences from 1964 through 2025. Spawning-biomass differences remain within one percent and equal minus 0.4 percent in 2025. Recruitment differences remain within four percent and equal minus 2.0 percent in 2025. Predicted-catch differences remain within 0.1 percent."
#| fig-width: 10
#| fig-height: 7
bts_age_comparison |>
  mutate(
    quantity = recode(
      quantity,
      spawning_biomass = "Spawning biomass",
      recruitment_age1 = "Age-1 recruitment",
      predicted_catch = "Predicted catch"
    )
  ) |>
  ggplot(aes(year, percent_difference)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_line(color = "#1f4e79", linewidth = 0.7) +
  geom_point(color = "#1f4e79", size = 1.1) +
  facet_wrap(vars(quantity), scales = "free_y", ncol = 1) +
  labs(x = "Year", y = "Difference from prior-age sensitivity (%)") +
  theme(legend.position = "none")
```

### Planned 2026 survey-data update {#sec-planned-2026-survey-data}

Data from the 2026 Acoustic-Trawl Survey (ATS), along with 2026 AVO data, will
be added to this report after processing and quality-control review are
complete. The update will identify any resulting changes to the relevant time
series, figures, tables, and model inputs.
