---
output: html
title: "Effective degrees of freedom (EDF)"
date: May 2025
editor_options: 
  chunk_output_type: console
bibliography: references.bib
editor: 
  markdown: 
    wrap: 90
---

```{r setup}
#| echo: false
#| label: setup
#| message: false
#| warning: false
#| 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(ggridges)
library(ebswp)
library(tidyverse)
library(patchwork)
library(viridis)
library(ggsidekick)
theme_set(theme_sleek())
.THEME=theme_sleek()
.OVERLAY=TRUE
do_edf <- FALSE
#plot_ssb(loo_lst)
```


# Developments on estimating the effective degrees of freedom

In @zheng2024 they present a new method for estimating the effective degrees of freedom
(EDF) for general model applications. They note that "...hierarchical models can express
ecological dynamics using a combination of fixed and random effects, and measurement of
their complexity requires estimating how much random effects are shrunk toward a shared
mean. Estimating EDF is helpful to (1) penalize complexity during model selection and (2)
to improve understanding of model behavior. ” Specifically, with application to common
stock assessment problems we wish to evaluate the relationship between the information
content of the data and the level of model complexity. See the paper for details (equaion
6). The conceptual steps to implement are based on the availability of the Hessian matrix.
The Hessian matrix is a square matrix of second-order partial derivatives of a
scalar-valued function. In this case, the Hessian is used to approximate the curvature of
the log-likelihood function with respect to the model parameters. The diagonal elements of
the Hessian matrix represent the second partial derivatives of the log-likelihood with
respect to each parameter, and they can be used to estimate the EDF for each parameter.
The first step is to estimate the Hessian from the full log-posterior (data and priors/penalties). Then, we map
out the effect of the data on the log-likelihood by only the effect of the
priors/penalties. This gives insight on the non-data impact on log-likelihood. For the
pollock model, we added a command-line argument "`-do_nodata`".

@tbl-edf1

@tbl-edf2

@tbl-edf3


```{r nodata}
#| output: asis
#| echo: false

#library(adnuts)
library(here)
if (do_edf){
  #---Redo of Cole's get hessian routine to get files----
  getADMBHess <- function(filename = "admodel.hes") {
    if (!file.exists(filename)) {
      stop(paste0("file not found: ", filename))
    }
    f <- file(filename, "rb")
    on.exit(close(f))
    num.pars <- readBin(f, "integer", 1)
    hes.vec <- readBin(f, "numeric", num.pars^2)
    hes <- matrix(hes.vec, ncol = num.pars, nrow = num.pars)
    hybrid_bounded_flag <- readBin(f, "integer", 1)
    scale <- readBin(f, "numeric", num.pars)
    return(hes)
  }
  Get_EDF <- function(rundir, redo=FALSE) {
    if(redo){
      file.remove(here::here(rundir, "h1.hes")) 
      file.remove(here::here(rundir, "h0.hes")) 
    }
    if (file.exists(here::here(rundir, "h1.hes")) &  file.exists(here::here(rundir, "h0.hes")) ) {
    }else{
      setwd(here::here(rundir))
      system("pm ") 
      system("cp admodel.hes h1.hes")
      
      system("pm -nodata -binp pm.bar -phase 22 -maxfn 0")
      system("cp admodel.hes h0.hes")
    }
    ## Read in the Hessians
    h1 <- getADMBHess(filename=here::here(rundir, "h1.hes"))
    h0 <- getADMBHess(filename=here::here(rundir, "h0.hes"))
    ## Create a vector of indices where fixed-effect parameters----
    fit <- read_fit( here("runs/lastyrdb/pm"))
    par.names <- fit$names[1:fit$nopar]
    names(fit)
    nll <- fit$nlogl
    #pars <- adnuts:::.read_mle_fit("pm", here::here(rundir))
    indx <- h1[1, ] ;
    indx <- as.logical(indx)
    indx[] <- FALSE
    ## Read in the parameter names----
    indx[] <- (grepl("dev",par.names))
    ## Total number of fixed-effect parameters----
    sum(indx)
    ## Get the submatrices----
    h1s <- h1[indx, indx]
    h0s <- h0[indx, indx]
    ## Compute the negative EDF----
    negEDF <- Matrix::diag(Matrix::solve(h1s, h0s))
    ## Put in  a dataframe because jim hates tapply----
    df <- data.frame(negEDF = negEDF, par_names = sub("\\[.*", "", par.names[indx]))
    tab <- df |> group_by(par_names) |> summarize(N = n(), edf = N - sum(negEDF))
    tab <- rbind(tab,df |> summarize(par_names="Total",N = n(), edf = N - sum(negEDF)))
    tab <- rbind(tab,df |> summarise(par_names="AIC",  N = " ", edf = 2*nll + 2*(n() - sum(negEDF)) ) )
    return(tab)
  }
  
  t1<- Get_EDF(rundir = here::here(rundir="runs/lastyrdbae/")) |> mutate(Model="Base")  #|> 
  t2<- Get_EDF(rundir = here::here(rundir="runs/ftnir1/"), redo=FALSE)   |> mutate(Model="FTNIR1")  
  t3<- Get_EDF(rundir = here::here(rundir="runs/ftnir2/"), redo=FALSE)   |> mutate(Model="FTNIR2")  
  
  p1 <- t1
  p2 <- Get_EDF(rundir = here::here(rundir="runs/separable/"), redo=FALSE) |> mutate(Model="Separable F")  #|> 
  p3 <- Get_EDF(rundir = here::here(rundir="runs/vpa/"),       redo=FALSE) |> mutate(Model="VPA-like")  #|> 
  p4 <- Get_EDF(rundir = here::here(rundir="runs/const0.1/"),  redo=FALSE) |> mutate(Model="Const-0.1")  #|> 
  p5 <- Get_EDF(rundir = here::here(rundir="runs/const0.05/"),  redo=FALSE) |> mutate(Model="Const-0.05")  #|> 
  
  q2 <- Get_EDF(rundir = here::here(rundir="runs/rewt/"), redo=FALSE) |> mutate(Model="Francis re-wt")  #|> 
  # Save all variables into one file
  save(t1, t2, t3, p1, p2, p3, p4, p5, q2, file = here::here("edf_data.RData"))
}else{
  load(file = here::here("edf_data.RData"))
}

edf_table <- function(df) {
  n_check <- df %>%
    group_by(par_names) %>%
    summarize(n_unique = n_distinct(N))
  
  # If all n_unique values are 1, then proceed with the transformation
  df_wide <- df %>%
    # Get one copy of N per par_names
    group_by(par_names) %>%
    mutate(N = first(N)) %>%
    ungroup() %>%
    # Convert to wide format
    pivot_wider(
      id_cols = c(par_names, N),
      names_from = Model,
      values_from = edf
    )
  
  df_wide |>  gt::gt() |> 
    # Format all numeric columns except N to one decimal place
    gt::fmt_number(
      columns = -c(par_names, N),  # All columns except par_names and N
      decimals = 0
    ) %>%
    # Keep N as integers (no decimals)
    gt::fmt_number(
      columns = N,
      decimals = 0
    ) %>%
    # Add table header
    gt::tab_header(
      title = "EDF Values by Model and Parameter"
    )
}

```

```{r edf1}
#| echo: false
#| label: tbl-edf1
#| tbl-cap: "Effective degrees of freedom (EDF) for the different models. The models are the base model (Base), the FT-NIRs age estimates with fleet-aggregated data (FTNIR1), and the FT-NIRs age estimates with fleet-specific data (FTNIR2)."
edf_table (df=rbind(t1, q2))
#gt::gt(tab) |> gt::fmt_number(columns = 3, decimals = 1)
```
```{r edf2}
#| echo: false
#| label: tbl-edf2
#| tbl-cap: "Effective degrees of freedom (EDF) for the different models. The models are the base model (Base), the FT-NIRs age estimates with fleet-aggregated data (FTNIR1), and the FT-NIRs age estimates with fleet-specific data (FTNIR2)."
edf_table (df=rbind(p1,p5,p4,p3))
```
```{r edf3}
#| echo: false
#| label: tbl-edf3
#| tbl-cap: "Effective degrees of freedom (EDF) for the different models. The models are the base model (Base), the FT-NIRs age estimates with fleet-aggregated data (FTNIR1), and the FT-NIRs age estimates with fleet-specific data (FTNIR2)."
edf_table (df=rbind(t1, t2, t3))
```
