---
output: html
title: "Research and developments on the eastern Bering Sea walleye pollock stock assessment"
subtitle: "September 2025,  Plan Team Draft"
author: "Jim Ianelli, Carey McGilliard, Sophia Wassermann"
lightbox: true
engine: knitr
date: now
date-format: "YYYY-MM-DD HH:mm"
css: mystyle.css
format:
  html:
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            inlineMath: [['\\(', '\\)'], ['$', '$']],
            displayMath: [['\\[', '\\]'], ['$$', '$$']],
            processEscapes: true
          }
        };
        </script>
    theme:
      light: [cosmo] 
      dark: [darkly]
    code-copy: true
    code-overflow: wrap
    toc: true
    toc-title: 'Contents'
    toc-location: left
    fig-cap-location: margin
    tbl-cap-location: margin
    embed-resources: true
    #numbering: false
    #number-sections: false
    code-summary: "Show the code"
    code-fold: true
  pdf:
    documentclass: scrreprt
    papersize: letter
    #numbering: false
    #number-sections: false
    header-includes:
      - \counterwithout{figure}{chapter}
      - \usepackage{fancyhdr}
      - \pagestyle{fancy}
      - \fancyhf{}
      - \lhead{May}
      - \rhead{2025}
      - \chead{--Draft--}
      - \rfoot{\thepage}
      - \lfoot{EBS pollock assessment discussion paper }
      
editor_options: 
  chunk_output_type: console
bibliography: references.bib
editor: 
  markdown: 
    wrap: 90
---

```{r startup, echo=FALSE, warnings=FALSE, message=FALSE, eval=TRUE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(plotly)
library(corrplot)
library(r4ss)
library(ggridges)
library(ebswp)
library(tidyverse)
library(patchwork)
library(gganimate)
library(ggridges)
library(viridis)
library(gt)
library(knitr)
library(ggsidekick)
library(scales)
# Set ggplot theme
# devtools::install_github("seananderson/ggsidekick")
thisyr=2024
theme_set(theme_sleek())
.THEME <- theme_sleek()
.OVERLAY <- TRUE
options(warn = -1)
do_sigmaR_profile <- 0
do_edf <- 0
do_ft <<- FALSE
do_abc_sel <- 0
do_yearloop <- 0
do_srrlst <- 0
do_selsens <- 0
do_loo <- 0 #this also turns on/off making the mmat rds
# F_t <- read_table(here::here("runs/lastyr/F40_t.rep"))
# names(F_t)
# F_t |> select(Year, F35, Fmsy) |> pivot_longer(  cols = c(F35, Fmsy), names_to = "Type", values_to = "F" ) |>
# ggplot(aes(x=Year, y=F, color=Type )) + geom_point() + geom_line() + ggthemes::theme_few() + ylim(c(0,NA))

Plot_Sel <- function(sel, fage = 1, lage = 13, nages = 15, styr = 1964, endyr = 2024) {
  sdf <- pivot_longer(sel, names_to = "age", values_to = "sel", cols = 2:(nages + 1)) |>
    filter(Year >= styr, Year <= endyr) |>
    mutate(age = as.numeric(age))
  p1 <- sdf |> ggplot(
    aes(x = age, y = as.factor(Year), height = sel)
  ) +
    geom_density_ridges(
      stat = "identity", scale = 2.8,
      alpha = .3, fill = "goldenrod", color = "tan"
    ) +
    ggthemes::theme_few() +
    ylab("Year") +
    xlab("Age (years)") +
    scale_x_continuous(limits = c(fage, lage), breaks = fage:lage) +
    scale_y_discrete(limits = rev(levels(as.factor(sdf$Year)))) +
    facet_grid(. ~ source)
  return(p1)
}

loadup <- FALSE
# loadup<-TRUE
if (loadup) {
  source(here::here("R", "GetResults.R"))
  source(here::here("runs/spock", "FitSpock.R"))
  # source(here::here("runs","spock", "FitSpock.R"))
  spock <- spock_obj(src = "RTMB")
  
  #spock <- list(sel = sel, ts = ts, obj = m1)
  # names(spock)
  r1 <- SS_obj(SS_output(dir = here::here("runs", "ss", "base"), verbose = FALSE), src = "SS3 base")
  r2 <- SS_obj(SS_output(dir = here::here("runs", "ss", "tight"), verbose = FALSE), src = "SS3 Constrained")
  r3 <- SS_obj(SS_output(dir = here::here("runs", "ss", "flex"), verbose = FALSE), src = "SS3 Flexible")
  c1 <- cea_obj(cea_run = pollock_base, src = "CEA, FE ")
  #c2 <- cea_obj(cea_run = fm0_re1, src = "CEA, R_RE")
  c2 <- cea_obj(cea_run = pollock_RE, src = "CEA, RE")
  # gp_run<-gp_obj()
  pm_run <- pm_obj()
  # pm_run2<-pm_obj(pm2, src="Pollock_model VPA-like")
  # load(here("SAM","poll23","run","model2.RData"))
  sam_run <- SAM_obj()
  # am_run<-AMAK_obj(run_dir=here("amak2","runs", "base"))
  # am_run2<-AMAK_obj(run_dir=here("amak2","runs", "3par"),src="AMAK-3par-logistic")
  # am_run3<-AMAK_obj(run_dir=here("amak2","runs", "cpue"),src="AMAK w CPUE", nind=6)
  # all_sel <- rbind(sam_run$sel,pm_run$sel,am_run$sel,ss_run$sel,gp_run$sel)
  # all_sel <- rbind(pm_run$sel,r1$sel,r2$sel,r3$sel)#,gp_run$sel)

  # all_ts <- rbind( sam_run$ts,pm_run$ts,am_run$ts,ss_run$ts,gp_run$ts)
  save.image(file = here::here("compares.Rdata"))
} else {
  load(here::here("compares.Rdata"))
  spock$sel <- spock$sel %>%
    rowwise() %>%
    mutate(across(`1`:`15`, ~ . / max(c_across(`1`:`15`)))) %>%
    ungroup()
#  spock$sel 
}
# Plot_Sel() + ggthemes::theme_few(base_size = 9)
```

# Summary

The work presented below was based on the same data sets used in the 2024 assessment. 
We broke the work into two parts, first is the discussion paper presented to the
Center for Independent Experts (CIE) review panel prepared in May 2025 (link here). 
FT-NIR data with the traditional data sources (BTS and ATS). The results of these
comparisons are presented below. 
Pre and post CIE review...

The model fits to the data are presented in the tables at
the end of the document. The first section steers towards the R library used for
processing model results for EBS pollock. The second section presents the comparisons of
the FT-NIR data with the traditional data sources. The third section compares results of
different software platforms (e.g., SS3, CEA, and AMAK) for the assessment. The fourth
section shows results comparing newer model-based estimates from survey data.


#  Response to 2025 CIE Review of EBS Walleye Pollock Assessment

This section outlines the response  to the findings and recommendations of 
the 2025 Center for Independent Experts 
(CIE) review of the Eastern Bering Sea (EBS) walleye pollock stock assessment.
The final peer-reviewed reports from the three reviewers are provided at this link
https://www.st.nmfs.noaa.gov/science-quality-assurance/cie-peer-reviews/cie-review-2025 . These documents summarize their findings and recommendations in detail.

Since the review: 

- ✅ The operational ADMB-based model has been fully ported to **RTMB**, with verification that outputs match the prior version. 

- 🔄 Initial steps are underway to address **Russian catch** uncertainty through scenario modeling based on
available data.

### 1. Model Platform and Estimation Framework

The full ADMB model has been implemented in RTMB, enabling: - Random walk selectivity and recruitment deviations - Improved error propagation via marginal likelihood - Mohn’s rho, retrospective peels, and full simulation integration

🟢 **Status**: Complete\
📝 See attached illustration chunk for output comparison (add figures/tables here).


### 2. Russian Removals and Stock Exchange

Russian catch has not historically been included in the assessment. All three reviewers noted this omission and recommended: - Conducting **sensitivity runs** - Quantifying the **scale and direction of transboundary movement** - Developing **collaborative or joint modeling strategies**

🟡 **Status**: In progress\
📌 Ongoing work involves sensitivity runs with assumed Russian removals of 20–45% of the EBS US catch, guided by UCSL (2024) and reviewer recommendations.


### 3. Process Uncertainty, Selectivity, and Growth

Key issues raised: - Random walk selectivity penalties are currently fixed. Reviewers suggested estimating process variances via marginal likelihood (Anders). - Weight-at-age models could be expanded to include cohort and temperature effects (Yan). - Estimation of time-varying natural mortality should be explored cautiously (Yan).

⚪ **Status**: Partially addressed — EDF and penalty variance exploration will proceed after Russian catch integration.


### 4. Data Weighting and Compositional Likelihoods

Matt Cieri strongly recommended testing a **Dirichlet-multinomial (DM)** approach via Laplace approximation to better reflect age composition variance.

⚪ **Status**: Under review — current bootstrap-based effective N retained; DM implementation deferred until after RTMB stabilization.


### 5. Retrospective Performance

Mohn’s rho was computed and incorporated into RTMB outputs. Diagnostic plots are available for: - Spawning biomass (SSB) - Recruitment (R) - Fishing mortality (F)

✅ **Status**: Complete


### 6. Summary of Reviewer Recommendations and Status

```{r}
#| label: tab-cie-final-responses
#| tbl-cap: "CIE 2025 Recommendations and Response Summary"
tribble(
  ~Theme, ~`Lead Reviewer`, ~Recommendation, ~Status,
  "RTMB transition", "All", "Port model from ADMB to RTMB", "✅ Completed",
  "Russian catch", "All", "Incorporate or test Russian removals", "🟡 In progress",
  "Selectivity complexity", "Nielsen", "Estimate process variances via marginal likelihood", "⚪ Planned",
  "Natural mortality", "Jiao", "Explore M variation and CEATTLE integration", "⚪ Conceptual",
  "Growth & WAA", "Jiao", "Model cohort and temperature effects on WAA", "⚪ Deferred",
  "Data weighting", "Cieri", "Test Dirichlet-multinomial likelihoods", "⚪ Deferred",
  "Retrospective diagnostics", "Nielsen", "Use Mohn's rho and integrate peels", "✅ Completed"
) %>% 
  gt::gt()
```




# ADMB → RTMB Code 

subtitle: "Tracking translation details while bridging ADMB to RTMB"

This standalone appendix tracks items to verify or refactor as the model is bridged from **ADMB** to **RTMB**. Each entry records the ADMB behavior, intended RTMB behavior, priority, and links/tests.

## Model implementation

The RTMB model code is implemented in the `R/` directory, with the main entry point 
in `R/Rpm.R`. 


## Comparison of ADMB and RTMB by components

The initial steps to bridge ADMB to RTMB involve identifying key components and their behaviors. The following table summarizes the components, their ADMB behavior, intended RTMB behavior, and current status (@tbl-comparison).

```{r}
#| label: tbl-comparison
#| tbl-cap: "Comparison of the rtmb code (`RPM`) and the ADMB code results."
#| echo: false
#| results: asis
#| message: false
#| warning: false
#| error: false
source(here::here("R/config.R"))
data$return_nll_only <- FALSE ; 
gt_compare_table2(rpm(parms), pm, tolerance = 1e-5, sort_by_diff=FALSE) 
data$return_nll_only <- TRUE
```

After some initial testing we were able to take the parameter estimates from the ADMB model and
show that the gradients were very close to zero and the negative log-likelihood values matched.

```{r}
#| results: asis
#| message: false
#| warning: false
#| error: false
#| label: tpl-gradient
#| tbl-cap: "Top 20 parameters with largest gradients (absolute value) from RTMB using ADMB parameter estimates."
#| echo: false
#| 
data.frame(par = names(obj$par), gr = abs(as.numeric(obj$gr())), gr_orig = as.numeric(obj$gr())) |> 
   arrange(desc(gr))  |> select(Parameter=par,gradient=gr_orig) |> head(20) |> gt::gt()
```


<!-- ## Tuning -->


<!-- The following code chunk generates a plot comparing the log values of two series (`r1` and `r2`) from the `df` data frame. The data is reshaped to a long format for easier plotting with `ggplot2`. -->
<!-- So in config.R file I can turn off parameter sets and tell which ones are poorly estimated or ones that should be turned off completely. I may want to turn off creating the map from zero value parameters and I still need to work on getting the deviation parameters to work correctly with the right penalty. -->

### Reparameterizing dev-vectors

The ADMB code has lots of cases where say $y_i$ is a function of a parameter $\theta$ and a deviation $\epsilon_i$ such that 
$y_i = \theta + \epsilon_i$. In ADMB, the deviations are estimated as a vector of parameters
with a constraint to have mean zero., 
In RTMB, it seems best to reparameterize them as a simple vector of parameters (say as $\theta_i$) 
and then do the likelihood parts as deviations from the mean. 

So far we've done this for the following parameters:

  * `log_avgrec` (average recruitment)
  * `log_avg_F` (average fishing mortality)
  * `log_slp_bts` (slope of the selectivity for the bottom trawl)
  * `log_a50_bts` (age-50% selected for the bottom trawl)
 
The results from feeding the MLE parameter estimates from ADMB
into the RTMB code gave exactly the same estimates of spawning stock biomass (SSB) and
age 1 recruits (@fig-SSB_recruits; @fig-diffs).
  
```{r}
#| fig-cap: "SSB and age-1 recruits from RTMB and ADMB, dashed lines are standard deviations (x10). Both ADMB and RTMB results are overlain."
#| echo: false
#| label: fig-SSB_recruits
#| message: false
#| results: hide
#| warning: false
#| fig-width: 6
#| fig-height: 8


#Top######################################
source(here::here("R/config.R"))
df_grad <- data.frame(par = names(obj$par), gr = abs(as.numeric(obj$gr())), gr_orig = as.numeric(obj$gr())) |> 
arrange(desc(gr))  |> select(Parameter=par,gradient=gr_orig) |> head(10)

fit <- nlminb(obj$par, obj$fn, obj$gr); 
# cest <- check_estimability(obj);cest$BadParams[cest$BadParams$Param_check=="Bad",]
sdr <- sdreport(obj)
rtmb_df <- tibble( name = names(sdr$value), sd   = sdr$sd )
rtmb_df |> filter(sd==0) |> print(n=Inf)
vals <- sdr$value
sds  <- sdr$sd

df <- rbind(data.frame(
  name  = names(vals),
  value = as.numeric(vals),
  std.dev = as.numeric(sds),
  model = "RTMB"
) , #|> filter(name=="SSB"), 
data.frame(
  read_table(here::here("runs/rtmb", "pm.std")) |> 
    filter(name == "SSB" | name == "pred_rec") |>  
    transmute( name = ifelse(name == "SSB", "SSB", "Recruits"), 
      value,  std.dev,  model = "ADMB" )
)) |> 
  mutate( name = ifelse(name == "recruitment", "Recruits", name) )
df1 <- df |> filter(name %in% c("SSB", "Recruits"))
df1$year <- rep(1964:2024,4) 
p1<- df1 |> filter(name=="SSB") |>   ggplot(aes(x=year,y=value,color=model)) +
  geom_line(linetype=2) +
  geom_line(aes(x=year,y=std.dev*10)) +
  labs(title = "RTMB vs ADMB SSB Standard Deviations", x = "Year ", y = "SSB, Standard Deviation x 10") +
theme_minimal() 
p2<- df1 |> filter(name!="SSB") |>   ggplot(aes(x=year,y=value,color=model)) +
  geom_line(linetype=2) +
  geom_line(aes(x=year,y=std.dev*10)) +
  labs(title = "RTMB vs ADMB recruitment Standard Deviations", x = "Year ", y = "recruits, Standard Deviation x 10") +
  theme_minimal()
p1/p2

```

```{r}
#| fig-cap: "Relative differences between ADMB and RTMB estimates of SSB and age-1 recruits and asymptotic standard deviation approximations." 
#| echo: false
#| label: fig-diffs
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 9
plot_ssb_value_ratio <- create_value_ratio_plot(df1, variable_name = "SSB", 
                                                numerator_model = "RTMB", 
                                                denominator_model = "ADMB")
plot_ssb_stddev_ratio <- create_stddev_ratio_plot(df1, variable_name = "SSB", 
                                                  numerator_model = "RTMB", 
                                                  denominator_model = "ADMB")
rec_ratio <- create_value_ratio_plot(df1, variable_name = "Recruits", 
                                                numerator_model = "RTMB", 
                                                denominator_model = "ADMB")
rec_sd_ratio <- create_stddev_ratio_plot(df1, variable_name = "Recruits", 
                                                  numerator_model = "RTMB", 
                                                  denominator_model = "ADMB")
(plot_ssb_value_ratio)/ (plot_ssb_stddev_ratio)/
(rec_ratio)/ (rec_sd_ratio)
```




```{r}
#| label: tuning
#| echo: false
#| eval: false
#| results: asis
#| message: false
#| warning: false
#| error: false

# names(parms)
source(here::here("R/config.R"))
obj$fn()
r1 <- obj$report()
parms$log_avgrec<-parms$log_avgrec*.96
parms$log_rec_devs<-parms$log_rec_devs*.6
parms$log_avg_F<-parms$log_avg_F*1.26
parms$sel_devs_fsh<-parms$sel_devs_fsh*0.6
obj <- MakeADFun(rpm, parms, map = map_obj)
rpm(parms)
fit1 <- nlminb(obj$par, obj$fn, obj$gr )
obj$fn()
for (i in 1:4) fit1<- nlminb(fit1$par, obj$fn, obj$gr )
  fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r2 <- obj$report()
 for (i in 1:5) 
  fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r3 <- obj$report()

for (i in 1:5) fit1<- nlminb(fit1$par, obj$fn, obj$gr)
  fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r4 <- obj$report()
for (i in 1:5) fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r5 <- obj$report()
df <- (data.frame(
  `from ADMB`=(unlist(r1)), 
  run_5x=(unlist(r2)),
  run10x=(unlist(r3)),
  run15x=(unlist(r4)),
   run20x=(unlist(r5))
  ))
df_long <- df %>%
  tibble::rownames_to_column("name") %>%
  mutate(across(-c(name, from.ADMB), ~ .x - from.ADMB)) %>%  # Subtract first column from others
  select(-from.ADMB) %>%  # Remove the reference column
  pivot_longer(cols = -name, names_to = "series", values_to = "difference") %>%
  filter(abs(difference) >0, name != "wt_like", name != "SSB", !grepl("SSB", name))  # Filter if needed
(ggplot(df_long, aes(x = name, y = (difference), fill = series)) +
  geom_col(position = "dodge") +
  coord_flip() +  # Optional, if you want horizontal labels
  # scale_y_continuous(limits = c(-9.5, 9.5)) +
  scale_fill_viridis_d() +  # Viridis discrete colors
  theme_minimal() +
  labs(x = "NLL Term", y = "difference from ADMB (absolute neg. log-likelihood units)", fill = "Series"))
```

# Considerations for future assessments regarding neighboring Russian federation fisheries

As this year’s assessment moves forward, particular attention is warranted on how Russian
fisheries and assessments are considered alongside our own results. Pollock in the eastern and
western Bering Sea are generally recognized as a single biological stock, with seasonal exchange
across the U.S.–Russia boundary documented by acoustic studies. As noted in the CIE reviews, 
Russian removals are substantial with recent estimates representing on the order of 20–45% of the U.S. catch in
some years.  
<!-- [oai_citation:0‡2025_08 Cieri EBS pollock assessment review report.pdf](file-service://file-DtNxGFHwQffPxqjpkEBdub) -->
While examined in past assessments, the Russian catch magnitude should be re-evaluated
to examine how they might alter management advice. 
<!-- [oai_citation:1‡2025_08 Cieri EBS pollock assessment review report.pdf](file-service://file-DtNxGFHwQffPxqjpkEBdub) [oai_citation:2‡2025_08 Nielsen EBS pollock assessment review report.pdf](file-service://file-9nnvtLxwuaXaY5D5r5qhm4).  -->
As a step in beginning to make a more detailed evaluation, we compared the Russian stock assessment
results presented as part of their 
Marine Stewardship Council (MSC) review panel with our model results (through 2024).
The following sections compare outcomes from the assessments presented to 


## Comparison of Russian and USA age‑structured assessment results 

```{r}
#| label: tbl-age_corr
#| tbl-cap: "Correlations between Russian and USA age classes, 1995-2024."
#| echo: false
#| message: false
#| warning: false
#| error: false 
# Load data
df <- read_csv(here::here("doc","data","nage.csv"))
russian_data <- df %>% filter(Source == "WBS") %>% select(Year, starts_with("Age"))
usa_data <- df %>% filter(Source == "EBS") %>% select(Year, starts_with("Age"))
# dim(usa_data);dim(russian_data)
# Create tidy data format
russian_long <- df |> filter(Source=="WBS") |> 
  pivot_longer(cols = starts_with("Age"), names_to = "Age", values_to = "Numbers") %>%
  mutate(Country = "Russia",
         Age = str_extract(Age, "\\d+") %>% as.numeric())

usa_long <- df |> filter(Source=="EBS") |> 
  pivot_longer(cols = starts_with("Age"), names_to = "Age", values_to = "Numbers") %>%
  mutate(Country = "USA",
         Age = str_extract(Age, "\\d+") %>% as.numeric())

# Combine datasets
combined_data <- bind_rows(russian_long, usa_long)

# 1. CORRELATION ANALYSIS
cat("=== CORRELATION ANALYSIS ===\n")

# Calculate correlations using wide format for easier computation
russian_wide <- combined_data %>% 
  filter(Country == "Russia") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

usa_wide <- combined_data %>% 
  filter(Country == "USA") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

# Age-specific correlations
age_correlations <- map_dbl(3:12, ~cor(russian_wide[[.x]], usa_wide[[.x]]))
names(age_correlations) <- paste0("Age", 1:10)

#cat("Correlations by Age Class:\n")
#print(round(age_correlations, 3)) 
# Overall correlation
all_russian <- russian_wide %>% select(-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
all_usa <- usa_wide %>% select(-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
overall_correlation <- cor(all_russian, all_usa, use = "complete.obs")
#cat("\nOverall correlation across all ages:", round(overall_correlation, 3), "\n")

# Create correlation dataframe for plotting
correlation_df <- tibble(
  Age = 1:10,
  Correlation = age_correlations
)
gt(correlation_df) %>%
  tab_header(
    title = "Correlations between Russian and USA Age Classes"
  ) %>%
  fmt_number(
    columns = vars(Correlation),
    decimals = 3
  )
```

Pollock ≈99% of catch; seabird & marine mammal mortality very low [@Artyukhin2023; @Burdin2021].
[@SharovJapp2023; @SharovJapp2024; @BulatovVasilev2023].

```{r}
#| eval: true
#| label: magnitude_comparison
#| tbl-cap: "Summary statistics of magnitude differences between Russian and USA age classes, 1995-2024."
#| echo: false
#| message: false
#| warning: false
#| error: false
#| results: asis
#| 
# 2. MAGNITUDE COMPARISON
# print("\n=== MAGNITUDE COMPARISON ===")

# Total numbers by country and year
russian_totals <- rowSums(russian_data[,-1])
usa_totals <- rowSums(usa_data[,-1])

magnitude_comparison <- data.frame(
  Year = russian_data$Year,
  Russian_Total = russian_totals,
  USA_Total = usa_totals,
  Ratio_USA_to_Russian = usa_totals / russian_totals
)

#print("Summary of magnitude differences:")
mn_russ_tot <- (paste("mean Russian total:", round(mean(russian_totals), 0)))
mn_usa_tot  <- (paste("mean USA total:", round(mean(usa_totals), 0)))
mn_ratio <- print(paste("mean ratio (USA/Russian):", round(mean(magnitude_comparison$Ratio_USA_to_Russian), 2)))
md_ratio <- print(paste("median ratio (USA/Russian):", round(median(magnitude_comparison$Ratio_USA_to_Russian), 2)))

# Age-specific magnitude ratios
# age_ratios <- sapply(2:11, function(i) {
#   mean(usa_data[,i] / russian_data[,i], na.rm = TRUE)
# })
# names(age_ratios) <- paste0("Age", 1:10)
# 3. GGPLOT VISUALIZATIONS
```

```{r}
#| eval: true
#| label: fig-plots_comparison
#| fig-cap: "Various comparisons between Russian and USA age-structured survey results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8

totals_data <- combined_data %>%
  group_by(Year, Country) %>%
  summarise(Total = sum(Numbers, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Country, values_from = Total) %>%
  mutate(Ratio_USA_to_Russian = USA / Russia)
# Plot 1: Total numbers over time
p1 <- totals_data %>%
  select(Year, Russia, USA) %>%
  pivot_longer(cols = c(Russia, USA), names_to = "Country", values_to = "Total") %>%
  ggplot(aes(x = Year, y = Total, color = Country)) +
  geom_line(size = 1.2) +
  geom_point(alpha = 0.7) +
  scale_y_continuous(labels = comma_format(), limits = c(0,NA)) +
  scale_color_manual(values = c("Russia" = "#E31A1C", "USA" = "#1F78B4")) +
  labs(title = "Total Numbers at Age Over Time",
       subtitle = "Comparison between Russian and USA surveys",
       x = "Year", y = "Total Numbers (thousands)",
       color = "Country") +
  theme_minimal() + ylim(c(0,NA)) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "top")

# Plot 2: Age 1 comparison over time
p2 <- combined_data %>%
  filter(Age == 1) %>%
  ggplot(aes(x = Year, y = Numbers, color = Country)) +
  geom_line(size = 1.2) +
  geom_point(alpha = 0.7) +
  scale_y_continuous(labels = comma_format(), limits=c(0,NA)) +
  scale_color_manual(values = c("Russia" = "#E31A1C", "USA" = "#1F78B4")) +
  labs(title = "Age 1 Numbers Over Time",
       x = "Year", y = "Numbers (thousands)",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "top")

# Plot 3: USA/Russian ratio over time
p3 <- totals_data %>%
  ggplot(aes(x = Year, y = Ratio_USA_to_Russian)) +
  geom_line(size = 1.2, color = "#FF7F00") +
  geom_point(alpha = 0.7, color = "#FF7F00") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50", alpha = 0.7) +
  labs(title = "USA/Russian Total Ratio Over Time",
       x = "Year", y = "Ratio (USA/Russian)") +ylim(c(0,NA)) +
  theme_minimal()

# Plot 4: Correlations by age class
p4 <- correlation_df %>%
  ggplot(aes(x = factor(Age), y = Correlation)) +
  geom_col(fill = "#33A02C", alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = c(0.5, -0.5), linetype = "dotted", alpha = 0.5, color = "red") +
  labs(title = "Correlations by Age Class",
       x = "Age Class", y = "Correlation Coefficient") +
  ylim(-1, 1) +
  theme_minimal()

# Plot 5: Age composition comparison (heatmap style)
p5 <- combined_data %>%
  group_by(Country, Age) %>%
  summarise(Mean_Numbers = mean(Numbers), .groups = "drop") %>%
  ggplot(aes(x = factor(Age), y = Country, fill = Mean_Numbers)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "#2166AC", 
                      trans = "log10", labels = comma_format()) +
  labs(title = "Mean Numbers by Age Class",
       subtitle = "Log-scale heatmap of average numbers across all years",
       x = "Age Class", y = "Country", fill = "Mean\nNumbers") +
  theme_minimal()

# Plot 6: Time series by age class (faceted)
p6 <- combined_data %>%
  filter(Age <= 6) %>%  # Focus on younger ages for clarity
  ggplot(aes(x = Year, y = Numbers, color = Country)) +
  geom_line(size = 1) +
  facet_wrap(~paste("Age", Age), scales = "free_y") +
  scale_y_continuous(labels = comma_format()) +
  scale_color_manual(values = c("Russia" = "#E31A1C", "USA" = "#1F78B4")) +
  labs(title = "Numbers at Age Over Time (Ages 1-6)",
       x = "Year", y = "Numbers (thousands)",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "top",
        strip.text = element_text(size = 10, face = "bold"))

# Combine plots using patchwork
combined_plot1 <- (p1 | p2) / (p3 | p4)
combined_plot2 <- p5 / p6
combined_plot1
```

```{r}
#| eval: true
#| label: fig-plots_comparison2
#| fig-cap: "Various comparisons between Russian and USA age-structured survey results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8
# Display plots
combined_plot2
```

```{r}
#| eval: true
#| label: fig-plots_comparison3
#| fig-cap: "Various comparisons between Russian and USA age-structured survey results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8
# 4. DETAILED STATISTICAL SUMMARY
cat("\n=== DETAILED STATISTICS ===\n")

# Summary statistics by country and age
stats_summary <- combined_data %>%
  group_by(Country, Age) %>%
  summarise(
    Mean = mean(Numbers),
    Median = median(Numbers),
    SD = sd(Numbers),
    CV = sd(Numbers) / mean(Numbers),
    Min = min(Numbers),
    Max = max(Numbers),
    .groups = "drop"
  ) %>%
  mutate(across(c(Mean, Median, SD, Min, Max), ~round(.x, 1)),
         CV = round(CV, 3))

cat("Summary statistics by country and age:\n")
stats_summary %>%
  arrange(Country, Age) %>%
  print(n = 20)

# Coefficient of variation comparison
cv_comparison <- stats_summary %>%
  select(Country, Age, CV) %>%
  pivot_wider(
    names_from = Country,
    values_from = CV,
    names_glue = "{Country}_CV"
  ) %>% arrange(Age)

cat("\nCoefficient of Variation by age:\n")
print(cv_comparison)

# Correlation analysis results
high_corr_ages <- correlation_df %>% filter(Correlation > 0.5) %>% pull(Age)
low_corr_ages <- correlation_df %>% filter(Correlation < 0.2) %>% pull(Age)

cat("\nAge classes with correlation > 0.5:", paste(paste0("Age", high_corr_ages), collapse = ", "), "\n")
cat("Age classes with correlation < 0.2:", paste(paste0("Age", low_corr_ages), collapse = ", "), "\n")

combined_data <- bind_rows(russian_long, usa_long)

# Calculate correlations using wide format for easier computation
russian_wide <- combined_data %>% 
  filter(Country == "Russia") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

usa_wide <- combined_data %>% 
  filter(Country == "USA") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

# Age-specific correlations
age_correlations <- map_dbl(3:12, ~cor(russian_wide[[.x]], usa_wide[[.x]]))
names(age_correlations) <- paste0("Age", 1:10)

# cat("Correlations by Age Class:\n")
# print(round(age_correlations, 3))

# Overall correlation
all_russian <- russian_wide %>% select(-Source,-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
all_usa <- usa_wide %>% select(-Source,-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
overall_correlation <- cor(all_russian, all_usa)
overall_corr <- paste0("Overall correlation across all ages:", round(overall_correlation, 3), "")
# Create correlation matrix for ggplot correlation plot
correlation_matrix_data <- combined_data %>% select(Year, Age, Numbers, Country) |> 
  pivot_wider(names_from = Country, values_from = Numbers) %>%
  select(-Year) %>%
  group_by(Age) %>%
  summarise(
    Correlation = cor(Russia, USA, use = "complete.obs"),
    .groups = "drop"
  )

```

Results comparing Russian and USA age-structured assessments 
`r print(overall_corr)`.


## Other items

The CIE noted in particular that some documentation of aspects of the ADMB code 
was incomplete. During the meetings, we began to more properly specify how the
selectivity penalties were being calculated and implementated.


### First‑difference selectivity penalty (temporal) {#test-fd-penalty-time}

We need the penalty used in ADMB where the **first year is compared to zero** (not to a previous estimated year). For a matrix of log‑selectivities ($\log s_{y,a}$) (years$(y=1,\ldots,T)$, ages $(a=1,\ldots,A)$, the penalty is:

$$
\mathcal{P}_{\text{FD,time}} \,=\, w \sum_{a=1}^{A}
\Big[
\big(\log s_{1,a}-0\big)^2
+ \sum_{y=2}^{T}\big(\log s_{y,a}-\log s_{y-1,a}\big)^2
\Big].
$$

#### RTMB/R implementation

```{r}
#| echo: true
# log_sel: matrix [T x A] of log-selectivities (year rows, age cols)
# w: scalar or length-A vector of weights
fd_penalty_time <- function(log_sel, w = 1) {
  stopifnot(is.matrix(log_sel))
  T <- nrow(log_sel); A <- ncol(log_sel)
  if (length(w) == 1) w <- rep(w, A)
  stopifnot(length(w) == A)

  diffs <- rbind(log_sel[1, , drop = FALSE] - 0,   # year 1 vs 0
                 apply(log_sel, 2, diff))          # y>=2 diffs
  sum(colSums((sqrt(w) * diffs)^2))
}
```

### Sanity check

```{r}
#| echo: true
set.seed(42)
T <- 5; A <- 3
L <- matrix(rnorm(T*A, sd=0.2), T, A); w <- c(1, 2, 0.5)

fd_loop <- function(L, w){
  acc <- 0
  for (a in seq_len(ncol(L))) {
    acc <- acc + w[a]*(L[1,a]-0)^2
    for (y in 2:nrow(L)) acc <- acc + w[a]*(L[y,a]-L[y-1,a])^2
  }
  acc
}

all.equal(fd_penalty_time(L, w), fd_loop(L, w))
```

### First‑difference selectivity penalty (age) {#test-fd-penalty-age}

For within-year smoothing over age with a **zero anchor at the minimum age**:

$$
\mathcal{P}_{\text{FD,age}} \,=\, w \sum_{y=1}^{T}
\Big[
\big(\log s_{y,a_{\min}}-0\big)^2
+ \sum_{a=a_{\min}+1}^{A}\big(\log s_{y,a}-\log s_{y,a-1}\big)^2
\Big].
$$

#### RTMB/R implementation

```{r}
#| echo: true
# log_sel: matrix [T x A] of log-selectivities (year rows, age cols)
# a_min: column index of minimum age within 'log_sel' (default = 1)
fd_penalty_age <- function(log_sel, w = 1, a_min = 1) {
  stopifnot(is.matrix(log_sel))
  T <- nrow(log_sel); A <- ncol(log_sel)
  stopifnot(a_min >= 1, a_min <= A)
  if (length(w) == 1) w <- rep(w, T)
  stopifnot(length(w) == T)

  # Build diffs across age with anchor at a_min (assumes a_min=1; if not, set zeros below a_min)
  diffs <- matrix(0, nrow = T, ncol = A)
  diffs[, a_min] <- log_sel[, a_min] - 0
  if (a_min < A) {
    for (a in (a_min+1):A) {
      diffs[, a] <- log_sel[, a] - log_sel[, a-1]
    }
  }
  sum(rowSums((sqrt(w) * diffs)^2))
}
```

## Reference: survey lognormal scaling (consistency check) {#test-survey-lognorm}

When porting lognormal survey likelihoods, confirm whether ADMB’s objective includes the normalizing constant and whether \(\sigma\) is on the **log** scale:

$\ell(\sigma) = -\sum_i \frac{(\log O_i - \log E_i)^2}{2\sigma^2} \; -\; n\log\sigma \; -\; \frac{n}{2}\log(2\pi).$

```{r}
#| echo: true
lognorm_nll <- function(obs, exp, sigma_log, include_const = TRUE) {
  r <- log(pmax(obs, .Machine$double.eps)) - log(pmax(exp, .Machine$double.eps))
  n <- length(r)
  nll <- sum(r^2) / (2 * sigma_log^2)
  if (include_const) nll <- nll + n*log(sigma_log) + n*0.5*log(2*pi)
  nll
}
```

# References
