---
output: html
title: "Research and developments on the eastern Bering Sea walleye pollock stock assessment"
subtitle: "September 2025,  Plan Team Draft"
author: "Jim Ianelli, Carey McGilliard, Sophia Wassermann"
lightbox: true
engine: knitr
date: now
date-format: "YYYY-MM-DD HH:mm"
css: mystyle.css
format:
  html:
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            inlineMath: [['\\(', '\\)'], ['$', '$']],
            displayMath: [['\\[', '\\]'], ['$$', '$$']],
            processEscapes: true
          }
        };
        </script>
    theme:
      light: [cosmo] 
      dark: [darkly]
    code-copy: true
    code-overflow: wrap
    toc: true
    toc-title: 'Contents'
    toc-location: left
    fig-cap-location: margin
    tbl-cap-location: margin
    embed-resources: true
    #numbering: false
    #number-sections: false
    code-summary: "Show the code"
    code-fold: true
  pdf:
    documentclass: scrreprt
    papersize: letter
    #numbering: false
    #number-sections: false
    header-includes:
      - \counterwithout{figure}{chapter}
      - \usepackage{fancyhdr}
      - \pagestyle{fancy}
      - \fancyhf{}
      - \lhead{May}
      - \rhead{2025}
      - \chead{--Draft--}
      - \rfoot{\thepage}
      - \lfoot{EBS pollock assessment discussion paper }
      
editor_options: 
  chunk_output_type: console
bibliography: references.bib
editor: 
  markdown: 
    wrap: 90
---

```{r startup, echo=FALSE, warnings=FALSE, message=FALSE, eval=TRUE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(plotly)
library(corrplot)
library(r4ss)
library(ggridges)
library(ebswp)
library(tidyverse)
library(patchwork)
library(gganimate)
library(ggridges)
library(viridis)
library(gt)
library(knitr)
library(ggsidekick)
library(scales)
# Set ggplot theme
# devtools::install_github("seananderson/ggsidekick")
thisyr=2024
theme_set(theme_sleek())
.THEME <- theme_sleek()
.OVERLAY <- TRUE
options(warn = -1)
do_sigmaR_profile <- 0
do_edf <- 0
do_ft <<- FALSE
do_abc_sel <- 0
do_yearloop <- 0
do_srrlst <- 0
do_selsens <- 0
do_loo <- 0 #this also turns on/off making the mmat rds
# F_t <- read_table(here::here("runs/lastyr/F40_t.rep"))
# names(F_t)
# F_t |> select(Year, F35, Fmsy) |> pivot_longer(  cols = c(F35, Fmsy), names_to = "Type", values_to = "F" ) |>
# ggplot(aes(x=Year, y=F, color=Type )) + geom_point() + geom_line() + ggthemes::theme_few() + ylim(c(0,NA))

Plot_Sel <- function(sel, fage = 1, lage = 13, nages = 15, styr = 1964, endyr = 2024) {
  sdf <- pivot_longer(sel, names_to = "age", values_to = "sel", cols = 2:(nages + 1)) |>
    filter(Year >= styr, Year <= endyr) |>
    mutate(age = as.numeric(age))
  p1 <- sdf |> ggplot(
    aes(x = age, y = as.factor(Year), height = sel)
  ) +
    geom_density_ridges(
      stat = "identity", scale = 2.8,
      alpha = .3, fill = "goldenrod", color = "tan"
    ) +
    ggthemes::theme_few() +
    ylab("Year") +
    xlab("Age (years)") +
    scale_x_continuous(limits = c(fage, lage), breaks = fage:lage) +
    scale_y_discrete(limits = rev(levels(as.factor(sdf$Year)))) +
    facet_grid(. ~ source)
  return(p1)
}

loadup <- FALSE
# loadup<-TRUE
if (loadup) {
  source(here::here("R", "GetResults.R"))
  source(here::here("runs/spock", "FitSpock.R"))
  # source(here::here("runs","spock", "FitSpock.R"))
  spock <- spock_obj(src = "RTMB")
  
  #spock <- list(sel = sel, ts = ts, obj = m1)
  # names(spock)
  r1 <- SS_obj(SS_output(dir = here::here("runs", "ss", "base"), verbose = FALSE), src = "SS3 base")
  r2 <- SS_obj(SS_output(dir = here::here("runs", "ss", "tight"), verbose = FALSE), src = "SS3 Constrained")
  r3 <- SS_obj(SS_output(dir = here::here("runs", "ss", "flex"), verbose = FALSE), src = "SS3 Flexible")
  c1 <- cea_obj(cea_run = pollock_base, src = "CEA, FE ")
  #c2 <- cea_obj(cea_run = fm0_re1, src = "CEA, R_RE")
  c2 <- cea_obj(cea_run = pollock_RE, src = "CEA, RE")
  # gp_run<-gp_obj()
  pm_run <- pm_obj()
  # pm_run2<-pm_obj(pm2, src="Pollock_model VPA-like")
  # load(here("SAM","poll23","run","model2.RData"))
  sam_run <- SAM_obj()
  # am_run<-AMAK_obj(run_dir=here("amak2","runs", "base"))
  # am_run2<-AMAK_obj(run_dir=here("amak2","runs", "3par"),src="AMAK-3par-logistic")
  # am_run3<-AMAK_obj(run_dir=here("amak2","runs", "cpue"),src="AMAK w CPUE", nind=6)
  # all_sel <- rbind(sam_run$sel,pm_run$sel,am_run$sel,ss_run$sel,gp_run$sel)
  # all_sel <- rbind(pm_run$sel,r1$sel,r2$sel,r3$sel)#,gp_run$sel)

  # all_ts <- rbind( sam_run$ts,pm_run$ts,am_run$ts,ss_run$ts,gp_run$ts)
  save.image(file = here::here("compares.Rdata"))
} else {
  load(here::here("compares.Rdata"))
  spock$sel <- spock$sel %>%
    rowwise() %>%
    mutate(across(`1`:`15`, ~ . / max(c_across(`1`:`15`)))) %>%
    ungroup()
#  spock$sel 
}
# Plot_Sel() + ggthemes::theme_few(base_size = 9)
```

# Summary

The work presented below was based on the same data sets used in the 2024 assessment. We
broke the work into two parts, first is the discussion paper presented to the Center for
Independent Experts (CIE) review panel prepared in May 2025 (link here). FT-NIR data with
the traditional data sources (BTS and ATS). The results of these comparisons are presented
below. Pre and post CIE review...

The model fits to the data are presented in the tables at the end of the document. The
first section steers towards the R library used for processing model results for EBS
pollock. The second section presents the comparisons of the FT-NIR data with the
traditional data sources. The third section compares results of different software
platforms (e.g., SS3, CEA, and AMAK) for the assessment. The fourth section shows results
comparing newer model-based estimates from survey data.

<!-- # Response to 2025 CIE Review of EBS Walleye Pollock Assessment -->

<!-- This section outlines the response to the findings and recommendations of the 2025 Center -->
<!-- for Independent Experts (CIE) review of the Eastern Bering Sea (EBS) walleye pollock stock -->
<!-- assessment. The final peer-reviewed reports from the three reviewers are provided at this -->
<!-- link -->
<!-- https://www.st.nmfs.noaa.gov/science-quality-assurance/cie-peer-reviews/cie-review-2025. -->
<!-- These documents summarize their findings and recommendations in detail. -->

<!-- Since the review: -->

<!-- -   ✅ The operational ADMB-based model has been fully ported to **RTMB**, with -->
<!--     verification that outputs match the prior version. -->

<!-- -   🔄 Initial steps are underway to address **Russian catch** uncertainty through -->
<!--     scenario modeling based on available data. -->

<!-- ### 1. Model Platform and Estimation Framework -->

<!-- The full ADMB model has been implemented in RTMB, enabling: - Random walk selectivity and -->
<!-- recruitment deviations - Improved error propagation via marginal likelihood - Mohn’s rho, -->
<!-- retrospective peels, and full simulation integration -->

<!-- 🟢 **Status**: Complete\ -->
<!-- 📝 See attached illustration chunk for output comparison (add figures/tables here). -->

<!-- ### 2. Russian Removals and Stock Exchange -->

<!-- Russian catch has not historically been included in the assessment. All three reviewers -->
<!-- noted this omission and recommended: - Conducting **sensitivity runs** - Quantifying the -->
<!-- **scale and direction of transboundary movement** - Developing **collaborative or joint -->
<!-- modeling strategies** -->

<!-- 🟡 **Status**: In progress\ -->
<!-- 📌 Ongoing work involves sensitivity runs with assumed Russian removals of 20–45% of the -->
<!-- EBS US catch, guided by UCSL (2024) and reviewer recommendations. -->

<!-- ### 3. Process Uncertainty, Selectivity, and Growth -->

<!-- Key issues raised: - Random walk selectivity penalties are currently fixed. Reviewers -->
<!-- suggested estimating process variances via marginal likelihood (Anders). - Weight-at-age -->
<!-- models could be expanded to include cohort and temperature effects (Yan). - Estimation of -->
<!-- time-varying natural mortality should be explored cautiously (Yan). -->

<!-- ⚪ **Status**: Partially addressed — EDF and penalty variance exploration will proceed -->
<!-- after Russian catch integration. -->

<!-- ### 4. Data Weighting and Compositional Likelihoods -->

<!-- Matt Cieri strongly recommended testing a **Dirichlet-multinomial (DM)** approach via -->
<!-- Laplace approximation to better reflect age composition variance. -->

<!-- ⚪ **Status**: Under review — current bootstrap-based effective N retained; DM -->
<!-- implementation deferred until after RTMB stabilization. -->

<!-- ### 5. Retrospective Performance -->

<!-- Mohn’s rho was computed and incorporated into RTMB outputs. Diagnostic plots are available -->
<!-- for: - Spawning biomass (SSB) - Recruitment (R) - Fishing mortality (F) -->

<!-- ✅ **Status**: Complete -->

<!-- ### 6. Summary of Reviewer Recommendations and Status -->

# Response to 2025 CIE Review of EBS Walleye Pollock Assessment

This section outlines the response to the findings and recommendations of the 2025 Center
for Independent Experts (CIE) review of the Eastern Bering Sea (EBS) walleye pollock stock
assessment. The final peer-reviewed reports from the three reviewers are provided at this
[link](https://www.st.nmfs.noaa.gov/science-quality-assurance/cie-peer-reviews/cie-review-2025).
These documents summarize their findings and recommendations.

Since the review, two major steps have been completed immediately: the operational
ADMB-based model has been fully ported to **RTMB**, with verification that outputs match
the prior version, and initial steps are underway to address **Russian catch** uncertainty
through scenario modeling.

One of the reviewers’ primary concerns was the continued reliance on ADMB. This has now
been resolved with the transition to RTMB, which enables use of random-walk selectivity
and recruitment deviations, improved error propagation through marginal likelihood, and
direct integration of retrospective diagnostics such as Mohn’s rho. This task is complete.

Another prominent theme was the absence of Russian removals from the assessment model. All
three reviewers emphasized the need to explore the magnitude of this omission, both in
terms of sensitivity runs and in better quantifying transboundary exchange. Work is in
progress, with planned runs assuming Russian removals of 20–45% of U.S. catch based on
(@SharovJapp2024) and reviewer guidance. These scenarios will provide bounds on the
possible scale of impacts before moving toward more collaborative or joint modeling
approaches.

With respect to process uncertainty, selectivity, and growth, the reviewers identified
several areas where additional flexibility could be valuable. Anders recommended
estimating process variances for random-walk selectivity penalties using marginal
likelihood, rather than fixing them. This aspect is possible within RCEATTLE and SPOCK
currently, and is now available in the RTMB version of the pollock model ("rpm"). Yan
noted that weight-at-age models could incorporate cohort and temperature effects, and also
suggested cautious exploration of time-varying natural mortality, potentially informed by
multispecies work such as CEATTLE. These issues have been partially addressed, and further
exploration of process error variances will follow after the Russian catch work is
integrated.

On data weighting and likelihood specification, Matt Cieri highlighted the potential
advantages of a **Dirichlet–multinomial (DM)** framework estimated via Laplace
approximation, as a more realistic representation of composition variance. For now, the
bootstrap-based effective sample size approach has been retained, with DM implementation
available in three different platforms (RCEATTLE, SPOCK, and the converted RTMB pollock
model. We anticipate examining this as an option in the coming year.

A consolidated summary of reviewer recommendations and their current status is provided in
Table @tbl-cie-final-responses.

```{r}
#| results: asis
#| message: false
#| warning: false
#| error: false
#| echo: false
#| label: tbl-cie-final-responses
#| tbl-cap: "CIE 2025 recommendations and response summary"
tribble(
  ~Theme, ~`Lead Reviewer`, ~Recommendation, ~Status,
  "RTMB transition", "All", "Port model from ADMB to RTMB", "✅ Completed",
  "Russian catch", "All", "Incorporate or test Russian removals", "🟡 In progress",
  "Selectivity complexity", "Nielsen", "Estimate process variances via marginal likelihood", "⚪ Planned",
  "Natural mortality", "Jiao", "Explore M variation and CEATTLE integration", "⚪ Conceptual",
  "Growth & WAA", "Jiao", "Model cohort and temperature effects on WAA", "⚪ Deferred",
  "Data weighting", "Cieri", "Test Dirichlet-multinomial likelihoods", "⚪  Available but deferred",
  "Retrospective ", "Nielsen", "Use Mohn's rho over shorter periods", "✅ Completed"
) %>% 
  gt::gt()
```

# Bridging ADMB to RTMB

The ADMB code used for the 2024 EBS pollock assessment has been fully ported to RTMB, with
verification that outputs match the prior version. This transition enables several
important enhancements, including the ability to estimate random-walk selectivity and
recruitment deviations, improved error propagation through marginal likelihood, and
greater transparency for code review and modifications. The original ADMB code has
several options for model configurations, including different selectivity models, data
inclusion, and process error structures. Only the base-case 2024 configuration was fully 
ported over to RTMB with some minor modifications as described below.

We changed the ADMB code as little as possible to 
enable it to be compatible and work within the RTMB framework. One change  was the use of
an `if` statement in the penalty for dome-shaped selectivity in the fishery. In the original ADMB code,
the penalty was invoked *if* the selectivity at one age was less than the selectivity at the
previous adjacent younger age. The modification to this was simply to penalize log-selectivities
of adjacent age groups older than age 6. 

The ADMB code also has several cases where say $y_i$ is a function of a parameter $\theta$ and
a deviation $\epsilon_i$ such that $y_i = \theta + \epsilon_i$. In ADMB, the deviations
are estimated as a vector of parameters with a constraint to have mean zero., In RTMB, it
seems best to reparameterize them as a simple vector of parameters (say as $\theta_i$) and
then do the likelihood parts as deviations from the mean.
To this end, we modified the following parameters which were parameterized as deviations from the mean
(e.g., `log_avgrec` (average recruitment), `log_avg_F` (average fishing
mortality), `log_slp_bts` (slope of the selectivity for the bottom trawl))
to be simply the vector of values.  These changes resulted in a slight change in the
pattern of spawning biomass and recruitment estimates among the 2024 ADMB model and the 2024
modified version of the ADMB model used for comparisons with the RTMB results (@fig-admbdiffs).

Once these minor modifications were made, the RTMB code was able to replicate the
results from feeding the MLE parameter estimates from ADMB into the RTMB code. This 
gave virtually the same estimates of spawning stock biomass (SSB) and age 1 recruits
(@fig-SSB_recruits; @fig-diffs).


```{r}
#| results: asis
#| message: false
#| warning: false
#| error: false
#| label: fig-admbdiffs
#| fig-cap: "Differences in SSB and recruits between the base ADMB used in 2024 and the modifications for bridging with RTMB. Note that these are slight modifications of the ADMB configurations."
#| fig-height: 8
#| fig-width: 8 
library(ebswp)
library(patchwork)
pm_orig <- read_rep(here::here("runs", "lastyr", "pm.rep")) # Read in the report file)
pm_mod  <- read_rep(here::here("runs", "rtmb", "pm.rep")) # Read in the report file)
sd_orig <- read_table(here::here("runs/lastyr", "pm.std")) |> filter(name=="pred_rec" )
sd_mod  <- read_table(here::here("runs/rtmb", "pm.std"))|> filter(name=="pred_rec" )
df <- rbind( data.frame( name  = "SSB",
  value = as.numeric(pm_orig$SSB[,2]),
  std.dev = as.numeric(pm_orig$SSB[,3]),
  model = "ADMB original" ) ,
  data.frame( name  = "SSB",
  value = as.numeric(pm_mod$SSB[,2]),
  std.dev = as.numeric(pm_mod$SSB[,3]),
  model = "ADMB Modified") )
df <- rbind(df, data.frame( name  = "Recruits",
    value = as.numeric(sd_orig$value),
    std.dev = as.numeric(sd_orig$std.dev),
    model = "ADMB original" ) ,
  data.frame( name  = "Recruits",
    value = as.numeric(sd_mod$value),
    std.dev = as.numeric(sd_mod$std.dev),
    model = "ADMB Modified") )
df$Year <- rep(1964:2024, 4)

p1 <- df |> filter(name=="SSB",Year>1977) |> 
  ggplot(aes(x=Year, y=value, color=model)) + geom_point() + geom_line(aes(group=model)) +
  geom_ribbon(aes(ymin=value-2*std.dev, ymax=value+2*std.dev, fill=model), alpha=0.2, color=NA) +
  ggthemes::theme_few() + ylab("SSB (t)") + xlab("Year") + 
  scale_y_continuous(labels = scales::comma, limits=c(0,NA)) + 
  scale_x_continuous(breaks=seq(1965,2025,5)) 

p2 <- df |> filter(name=="Recruits", Year>1977) |> 
  ggplot(aes(x=as.factor(Year), y=value, color=model)) + 
  geom_point(size=2, position=position_dodge(width=0.8)) + 
  geom_errorbar(aes(ymin=value-2*std.dev, ymax=value+2*std.dev), 
                alpha=0.8, width=0.8, position=position_dodge(width=0.8)) +
  ggthemes::theme_few() + 
  ylab("Age 1 Recruitment (thousands)") + 
  xlab("Year") + 
  scale_y_continuous(labels = scales::comma) + 
  expand_limits(y = 0) +
  scale_x_discrete(breaks=seq(1980,2025,5))
p1/p2
```

## RTMB Model implementation

The RTMB model code is implemented in the `R/` directory, with the main entry point in
`R/Rpm.R`. The initial steps to bridge ADMB to RTMB involve identifying key components and their
behaviors. We created a table function that was helpful to evaluate which components of the model
in RTMB tracked with the ADMB version. When the bridging aspect was acheieved there were
de minimus differences between the two model versions (@tbl-comparison).
Additionally, after some initial testing we were able to take the parameter estimates from the ADMB
model and show that the gradients were very close to zero and the negative log-likelihood
values matched (tbl-gradient).

```{r}
#| label: tbl-comparison
#| tbl-cap: "Comparison of the rtmb code (`RPM`) and the ADMB code results."
#| echo: false
#| results: asis
#| message: false
#| warning: false
#| error: false
source(here::here("R/config.R"))
data$return_nll_only <- FALSE ; 
gt_compare_table2(rpm(parms), pm, tolerance = 1e-5, sort_by_diff=FALSE) 
data$return_nll_only <- TRUE
```



```{r}
#| results: asis
#| message: false
#| warning: false
#| error: false
#| label: tbl-gradient
#| tbl-cap: "Top 20 parameters with largest gradients (absolute value) from RTMB using ADMB parameter estimates."
#| echo: false
#| 
data.frame(par = names(obj$par), gr = abs(as.numeric(obj$gr())), gr_orig = as.numeric(obj$gr())) |> 
   arrange(desc(gr))  |> select(Parameter=par,gradient=gr_orig) |> head(20) |> gt::gt()
```

<!-- ## Tuning -->

<!-- The following code chunk generates a plot comparing the log values of two series (`r1` and `r2`) from the `df` data frame. The data is reshaped to a long format for easier plotting with `ggplot2`. -->

<!-- So in config.R file I can turn off parameter sets and tell which ones are poorly estimated or ones that should be turned off completely. I may want to turn off creating the map from zero value parameters and I still need to work on getting the deviation parameters to work correctly with the right penalty. -->

```{r}
#| fig-cap: "SSB and age-1 recruits from RTMB and ADMB, dashed lines are standard deviations (x10). Both ADMB and RTMB results are overlain."
#| echo: false
#| label: fig-SSB_recruits
#| message: false
#| results: hide
#| warning: false
#| fig-width: 6
#| fig-height: 8


#Top######################################
source(here::here("R/config.R"))
df_grad <- data.frame(par = names(obj$par), gr = abs(as.numeric(obj$gr())), gr_orig = as.numeric(obj$gr())) |> 
arrange(desc(gr))  |> select(Parameter=par,gradient=gr_orig) |> head(10)

fit <- nlminb(obj$par, obj$fn, obj$gr); 
# cest <- check_estimability(obj);cest$BadParams[cest$BadParams$Param_check=="Bad",]
sdr <- sdreport(obj)
rtmb_df <- tibble( name = names(sdr$value), sd   = sdr$sd )
rtmb_df |> filter(sd==0) |> print(n=Inf)
vals <- sdr$value
sds  <- sdr$sd

df <- rbind(data.frame(
  name  = names(vals),
  value = as.numeric(vals),
  std.dev = as.numeric(sds),
  model = "RTMB"
) , #|> filter(name=="SSB"), 
data.frame(
  read_table(here::here("runs/rtmb", "pm.std")) |> 
    filter(name == "SSB" | name == "pred_rec") |>  
    transmute( name = ifelse(name == "SSB", "SSB", "Recruits"), 
      value,  std.dev,  model = "ADMB" )
)) |> 
  mutate( name = ifelse(name == "recruitment", "Recruits", name) )
df1 <- df |> filter(name %in% c("SSB", "Recruits"))
df1$year <- rep(1964:2024,4) 
p1<- df1 |> filter(name=="SSB") |>   ggplot(aes(x=year,y=value,color=model)) +
  geom_line(linetype=2) +
  geom_line(aes(x=year,y=std.dev*10)) +
  labs(title = "RTMB vs ADMB SSB Standard Deviations", x = "Year ", y = "SSB, Standard Deviation x 10") +
theme_minimal() 
p2<- df1 |> filter(name!="SSB") |>   ggplot(aes(x=year,y=value,color=model)) +
  geom_line(linetype=2) +
  geom_line(aes(x=year,y=std.dev*10)) +
  labs(title = "RTMB vs ADMB recruitment Standard Deviations", x = "Year ", y = "recruits, Standard Deviation x 10") +
  theme_minimal()
p1/p2

```

```{r}
#| fig-cap: "Relative differences between ADMB and RTMB estimates of SSB and age-1 recruits and asymptotic standard deviation approximations." 
#| echo: false
#| label: fig-diffs
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 9
plot_ssb_value_ratio <- create_value_ratio_plot(df1, variable_name = "SSB", 
                                                numerator_model = "RTMB", 
                                                denominator_model = "ADMB")
plot_ssb_stddev_ratio <- create_stddev_ratio_plot(df1, variable_name = "SSB", 
                                                  numerator_model = "RTMB", 
                                                  denominator_model = "ADMB")
rec_ratio <- create_value_ratio_plot(df1, variable_name = "Recruits", 
                                                numerator_model = "RTMB", 
                                                denominator_model = "ADMB")
rec_sd_ratio <- create_stddev_ratio_plot(df1, variable_name = "Recruits", 
                                                  numerator_model = "RTMB", 
                                                  denominator_model = "ADMB")
(plot_ssb_value_ratio)/ (plot_ssb_stddev_ratio)/
(rec_ratio)/ (rec_sd_ratio)
```

```{r}
#| label: tuning
#| echo: false
#| eval: false
#| results: asis
#| message: false
#| warning: false
#| error: false

# names(parms)
source(here::here("R/config.R"))
#obj$fn()
r1 <- obj$report()
parms$log_avgrec<-parms$log_avgrec*1.04
parms$log_rec_devs<-parms$log_rec_devs*.96
parms$log_avg_F<-parms$log_avg_F*1.06
parms$sel_devs_fsh<-parms$sel_devs_fsh*0.94
obj <- MakeADFun(rpm, parms, map = map_obj)
rpm(parms)
fit1 <- nlminb(obj$par, obj$fn, obj$gr )
r2 <- obj$report()
for (i in 1:4) fit1<- nlminb(fit1$par, obj$fn, obj$gr )
  fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r3 <- obj$report()
 for (i in 1:5) 
  fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r4 <- obj$report()

for (i in 1:5) fit1<- nlminb(fit1$par, obj$fn, obj$gr)
  fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r5 <- obj$report()
for (i in 1:5) fit1<- nlminb(fit1$par, obj$fn, obj$gr )
r5 <- obj$report()
df <- (data.frame(
  `from ADMB`=(unlist(r1)), 
  #run_5x=(unlist(r2)),
  #run10x=(unlist(r3)),
  run15x=(unlist(r4)),
   run20x=(unlist(r5))
  ))
df_long <- df %>%
  tibble::rownames_to_column("name") %>%
  mutate(across(-c(name, from.ADMB), ~ .x - from.ADMB)) %>%  # Subtract first column from others
  select(-from.ADMB) %>%  # Remove the reference column
  pivot_longer(cols = -name, names_to = "series", values_to = "difference") %>%
  filter(abs(difference) >0, name != "wt_like", name != "SSB", !grepl("SSB", name))  # Filter if needed
plotly::ggplotly(ggplot(df_long, aes(x = name, y = (difference), fill = series)) +
  geom_col(position = "dodge") +
  coord_flip() +  # Optional, if you want horizontal labels
  # scale_y_continuous(limits = c(-9.5, 9.5)) +
  scale_fill_viridis_d() +  # Viridis discrete colors
  theme_minimal() +
  labs(x = "NLL Term", y = "difference from ADMB (absolute neg. log-likelihood units)", fill = "Series"))
```

# Considerations for future assessments regarding neighboring Russian federation fisheries

As this year’s assessment moves forward, particular attention is warranted on how Russian
fisheries and assessments are considered alongside our own results. Pollock in the eastern
and western Bering Sea are generally recognized as a single biological stock, with
seasonal exchange across the U.S.–Russia boundary documented by acoustic studies. As noted
in the CIE reviews, Russian removals are substantial with recent estimates representing on
the order of 20–45% of the U.S. catch in some years.\
<!-- [oai_citation:0‡2025_08 Cieri EBS pollock assessment review report.pdf](file-service://file-DtNxGFHwQffPxqjpkEBdub) -->
While examined in past assessments, the Russian catch magnitude should be re-evaluated to
examine how they might alter management advice.
<!-- [oai_citation:1‡2025_08 Cieri EBS pollock assessment review report.pdf](file-service://file-DtNxGFHwQffPxqjpkEBdub) [oai_citation:2‡2025_08 Nielsen EBS pollock assessment review report.pdf](file-service://file-9nnvtLxwuaXaY5D5r5qhm4).  -->
As a step in beginning to make a more detailed evaluation, we compared the Russian stock
assessment results presented as part of their Marine Stewardship Council (MSC) review
panel with our model results (through 2024). The following sections compare outcomes from
the assessments presented to

## Comparison of Russian and USA age‑structured assessment results

### Comparison of Russian WBS and U.S. EBS Pollock Assessments

| Feature | Russian WBS Assessment | U.S. EBS Assessment |
|---------------------|------------------------------------|----------------------------------|
| **Primary model** | TISVPA catch-at-age model (Vasilyev 2005); Synthesis model retained as supporting tool | ADMB statistical catch-at-age model (pollock model), routinely updated |
| **Assessment frequency** | Annual assessment; fishery-independent surveys \~ every 2 years | Annual assessment; multiple fishery-independent surveys each year (BTS, ATS, AVO) |
| **Reference points (HCR)** | Since 2022: **Ftarget = Fmsy (or proxy)**; **Btrigger = 5th percentile of SSBmsy distribution** | Precautionary: **FABC \< Fmsy**, **B35%** proxy for SSBmsy; reference points set conservatively |
| **Management philosophy** | “Modern MSY” approach, aiming for maximum sustainable harvest without exceeding safe stock status; allows \~15–30% higher TACs vs. precautionary approach | Precautionary approach emphasizes risk reduction; TAC constrained by Council harvest rules |
| **Recruitment assumption** | Recruitment assumed independent of stock size, fluctuating randomly within historical distribution | Stock–recruit relationship incorporated (steepness parameter), with uncertainty reflected in projections |
| **Recent stock status** | 2023: SSB well above SSBtarget and SSBmsy proxy; F \< Ftarget and Flimit; stock increasing | 2024: SSB above B35% and Bmsy proxy; F well below reference points; stock stable to increasing |
| **Uncertainty notes** | Higher terminal year uncertainty in TISVPA compared to Synthesis; risk of overexploitation if stock declines under new HCR | Uncertainty routinely quantified (likelihood profiles, MCMC); precautionary buffers applied in catch advice |
| **Governance & compliance** | Managed by Federal Fisheries Agency (FFA) and CFMC; MSC certified since 2021; electronic logbook system in place | Managed under NPFMC/AFSC framework; extensive observer coverage; SAFE reports reviewed annually |

The contrasts between the Russian WBS and U.S. EBS pollock assessments highlight both
complementary perspectives and important methodological differences. Both systems conclude
that pollock biomass is currently above target reference levels and that fishing mortality
is below limits, reinforcing the perception of a healthy stock across the Bering Sea.
However, the Russian shift to a “modern MSY” framework, with $F_{target}$ set at $F_{msy}$
and $B_{trigger}$ at the lower 5th percentile of SSBmsy, allows higher catch levels but
introduces greater risk if recruitment declines. By contrast, the U.S. precautionary
approach constrains harvest below Fmsy and maintains conservative biomass thresholds,
prioritizing long-term risk reduction. These differences underscore the need for careful
interpretation when comparing results across assessments: while the overall trends align,
differences in model structure, reference points, and treatment of uncertainty can produce
divergent management advice even under similar stock conditions.

### Russian literature context: MSY vs. Precaution in WBS pollock

Bulatov & Vasilyev (2023) argue for moving from a precautionary HCR to a modern *MSY-based
rule* for WBS/Navarin and N. Okhotsk pollock. Their stochastic projections (using
*TISVPA*) resample historical recruitment (assumed independent of SSB) and set
$F_{target} \ge F_{MSY}$, with *MSY* $B_{trigger}$ taken as the 5th percentile of the SSB
distribution under $F_{MSY}$. Reported values include:

-   $F_{MSY}$ ≈ 0.75 yr⁻¹ (Navarin, WBS) and **0.60 yr⁻¹** (N. Okhotsk).\
-   $B_{trigger}$ ≈ 470 kt (Navarin) and **1.77 Mt** (N. Okhotsk).\
-   Expected catches **15–30% higher** than under the precautionary rule (2002–2022
    reconstructions), with large short-term gains in 2022 (e.g., Navarin TAC +\~60%).

**Implications for our EBS–WBS comparison:** the Russian MSY framing aligns with the
recent WBS HCR updates (F<sub>target</sub>=F<sub>MSY</sub>, low B-trigger), helping
explain higher Russian TAC recommendations when recruitment is strong. The approach is
intentionally less conservative and hinges on **weak SRR signal** assumptions; if
recruitment dips, risk tolerance differs from the EBS precautionary setup. We’ll reference
this when interpreting USA–Russia abundance ratios and management advice.

Datsky, Sheybak & Antonov (2022)\*\* provide a comprehensive synthesis of pollock
distribution, biology, stock status, and fishery dynamics across the Bering Sea. Drawing
on \>25 years of Russian survey and fishery data, they describe the **ecological
plasticity** of pollock, its broad bathymetric range (20–600 m), and the role of
**seasonal migrations** linking eastern and western Bering Sea populations.

Key points include:\
- **Population structure:** major groupings (eastern Bering Sea, Anadyr–Navarin, western
Bering Sea) have distinct spawning centers but mix during feeding migrations, particularly
in the northwestern Bering Sea.

-   **Stock trends:** the western Bering Sea stock declined to lows in the late
    1990s–early 2000s, but has shown **gradual recovery** since 2010. East Bering Sea and
    NW Bering Sea biomasses are currently at average long-term levels, with fluctuations
    strongly influenced by climate, ice, and prey availability.

-   **Fishery importance:** pollock accounts for \~70% of catches in Russian Bering Sea
    waters, dominated by midwater trawls, with recent landings of \~350–630 kt/yr.
    Seasonal patterns show \~90% of catches taken June–December.

**Implications for U.S.–Russia comparisons:** This work emphasizes that western and
northwestern pollock are not isolated; they are reinforced by migrations from the eastern
Bering Sea, meaning observed differences in assessment outputs (biomass, recruitment)
partly reflect shared dynamics across the basin as well as differences in survey intensity
and assessment methods.

```{r}
#| label: tbl-age_corr
#| tbl-cap: "Correlations between Russian and USA age classes, 1995-2024."
#| echo: false
#| message: false
#| warning: false
#| error: false 
# Load data
df <- read_csv(here::here("doc","data","nage.csv"))
russian_data <- df %>% filter(Source == "WBS") %>% select(Year, starts_with("Age"))
usa_data <- df %>% filter(Source == "EBS") %>% select(Year, starts_with("Age"))
# dim(usa_data);dim(russian_data)
# Create tidy data format
russian_long <- df |> filter(Source=="WBS") |> 
  pivot_longer(cols = starts_with("Age"), names_to = "Age", values_to = "Numbers") %>%
  mutate(Country = "Russia", Age = str_extract(Age, "\\d+") %>% as.numeric())

usa_long <- df |> filter(Source=="EBS") |> 
  pivot_longer(cols = starts_with("Age"), names_to = "Age", values_to = "Numbers") %>%
  mutate(Country = "USA", Age = str_extract(Age, "\\d+") %>% as.numeric())

# Combine datasets
combined_data <- bind_rows(russian_long, usa_long)

# 1. CORRELATION ANALYSIS
# cat("=== CORRELATION ANALYSIS ===\n")

# Calculate correlations using wide format for easier computation
russian_wide <- combined_data %>% 
  filter(Country == "Russia") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

usa_wide <- combined_data %>% 
  filter(Country == "USA") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

# Age-specific correlations
age_correlations <- map_dbl(3:12, ~cor(russian_wide[[.x]], usa_wide[[.x]]))
names(age_correlations) <- paste0("Age", 1:10)

#cat("Correlations by Age Class:\n")
#print(round(age_correlations, 3)) 
# Overall correlation
all_russian <- russian_wide %>% select(-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
all_usa <- usa_wide %>% select(-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
overall_correlation <- cor(all_russian, all_usa, use = "complete.obs")
#cat("\nOverall correlation across all ages:", round(overall_correlation, 3), "\n")

# Create correlation dataframe for plotting
correlation_df <- tibble( Age = 1:10, Correlation = age_correlations )
gt(correlation_df) %>% tab_header( title = "Correlations between Russian and USA Age Classes" ) %>%
  fmt_number( columns = vars(Correlation), decimals = 3 )
```

Pollock ≈99% of catch; seabird & marine mammal mortality very low @Artyukhin2023;
@Burdin2021\]. [@SharovJapp2023; @SharovJapp2024; @BulatovVasilev2023]. @Datsky2022,

```{r}
#| eval: true
#| label: tbl-magnitude_comparison
#| tbl-cap: "Summary statistics of magnitude differences between Russian and USA age classes, 1995-2024."
#| echo: false
#| message: false
#| warning: false
#| error: false
#| results: asis
#| 
# 2. MAGNITUDE COMPARISON
# print("\n=== MAGNITUDE COMPARISON ===")

# Total numbers by country and year
russian_totals <- rowSums(russian_data[,-1])
usa_totals <- rowSums(usa_data[,-1])

magnitude_comparison <- data.frame(
  Year = russian_data$Year,
  Russian_Total = russian_totals,
  USA_Total = usa_totals,
  Ratio_USA_to_Russian = usa_totals / russian_totals )

#print("Summary of magnitude differences:")
mn_russ_tot <- (paste("mean Russian total:", round(mean(russian_totals), 0)))
mn_usa_tot  <- (paste("mean USA total:", round(mean(usa_totals), 0)))
mn_ratio <- print(paste("mean ratio (USA/Russian):", round(mean(magnitude_comparison$Ratio_USA_to_Russian), 2)))
md_ratio <- print(paste("median ratio (USA/Russian):", round(median(magnitude_comparison$Ratio_USA_to_Russian), 2)))

# Age-specific magnitude ratios
# age_ratios <- sapply(2:11, function(i) {
#   mean(usa_data[,i] / russian_data[,i], na.rm = TRUE)
# })
# names(age_ratios) <- paste0("Age", 1:10)
# 3. GGPLOT VISUALIZATIONS
```

The comparison of U.S. and Russian pollock assessments shows that U.S. estimates of total
numbers-at-age are consistently higher than those from Russia, typically by a factor of
2–4, with the ratio peaking above 5 in the early 2010s (@fig-plots_comparison,
bottom-left). Both assessments show similar interannual patterns, particularly for age-1
recruits (top-right), suggesting some shared signals of strong year classes. Correlation
analysis by age class (bottom-right; @tbl-age_corr) confirms this alignment, with moderate
to strong positive correlations across ages 1–6. At older ages, correlations weaken and
even turn negative (age 10), reflecting differences in how the two assessments treat age
structure and fishery/survey data inputs. Overall, while the magnitude of estimates
differs substantially between the two countries, the consistent synchrony in recruitment
trends indicates that both assessments are capturing the same broad population dynamics,
albeit with systematic differences in scale.

```{r}
#| eval: true
#| label: fig-plots_comparison
#| fig-cap: "Various comparisons between Russian and USA age-structured survey results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8

totals_data <- combined_data %>% group_by(Year, Country) %>%
  summarise(Total = sum(Numbers, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Country, values_from = Total) %>% mutate(Ratio_USA_to_Russian = USA / Russia)
# Plot 1: Total numbers over time
p1 <- totals_data %>% select(Year, Russia, USA) %>%
  pivot_longer(cols = c(Russia, USA), names_to = "Country", values_to = "Total") %>%
  ggplot(aes(x = Year, y = Total, color = Country)) + geom_line(size = 1.2) +
  geom_point(alpha = 0.7) + scale_y_continuous(labels = comma_format(), limits = c(0,NA)) +
  scale_color_manual(values = c("Russia" = "#E31A1C", "USA" = "#1F78B4")) +
  labs(title = "Total Numbers at Age Over Time",
       subtitle = "Comparison between Russian and USA assessments",
       x = "Year", y = "Total Numbers (thousands)", color = "Country") +
  theme_minimal() + ylim(c(0,NA)) + theme(plot.title = element_text(size = 14, face = "bold"), legend.position = "top")

# Plot 2: Age 1 comparison over time
p2 <- combined_data %>% filter(Age == 1) %>% ggplot(aes(x = Year, y = Numbers, color = Country)) +
  geom_line(size = 1.2) + geom_point(alpha = 0.7) + scale_y_continuous(labels = comma_format(), limits=c(0,NA)) +
  scale_color_manual(values = c("Russia" = "#E31A1C", "USA" = "#1F78B4")) +
  labs(title = "Age 1 Numbers Over Time", x = "Year", y = "Numbers (thousands)", color = "Country") +
  theme_minimal() + theme(legend.position = "top")

# Plot 3: USA/Russian ratio over time
p3 <- totals_data %>%
  ggplot(aes(x = Year, y = Ratio_USA_to_Russian)) + geom_line(size = 1.2, color = "#FF7F00") +
  geom_point(alpha = 0.7, color = "#FF7F00") +theme_minimal() + 
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50", alpha = 0.7) +
  labs(title = "USA/Russian Total Ratio Over Time", x = "Year", y = "Ratio (USA/Russian)") +ylim(c(0,NA)) 

# Plot 4: Correlations by age class
p4 <- correlation_df %>%
  ggplot(aes(x = factor(Age), y = Correlation)) + geom_col(fill = "#33A02C", alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = c(0.5, -0.5), linetype = "dotted", alpha = 0.5, color = "red") +
  labs(title = "Correlations by Age Class", x = "Age Class", y = "Correlation Coefficient") +
  ylim(-1, 1) + theme_minimal()

# Plot 5: Age composition comparison (heatmap style)
p5 <- combined_data %>% group_by(Country, Age) %>%
  summarise(Mean_Numbers = mean(Numbers), .groups = "drop") %>%
  ggplot(aes(x = factor(Age), y = Country, fill = Mean_Numbers)) +
  geom_tile() + scale_fill_gradient2(low = "white", high = "#2166AC",  trans = "log10", labels = comma_format()) +
  labs(title = "Mean Numbers by Age Class", subtitle = "Log-scale heatmap of average numbers across all years",
       x = "Age Class", y = "Country", fill = "Mean\nNumbers") + theme_minimal()

# Plot 6: Time series by age class (faceted)
p6 <- combined_data %>% filter(Age <= 6) %>%  # Focus on younger ages for clarity
  ggplot(aes(x = Year, y = Numbers, color = Country)) + geom_line(size = 1) +
  facet_wrap(~paste("Age", Age), scales = "free_y") + scale_y_continuous(labels = comma_format()) +
  scale_color_manual(values = c("Russia" = "#E31A1C", "USA" = "#1F78B4")) +
  labs(title = "Numbers at Age Over Time (Ages 1-6)", x = "Year", y = "Numbers (thousands)", color = "Country") +
  theme_minimal() + theme(legend.position = "top", strip.text = element_text(size = 10, face = "bold"))

# Combine plots using patchwork
(p1 | p2) / (p3 | p4)
```

```{r}
#| eval: true
#| label: fig-plots_comparison2
#| fig-cap: "Various comparisons between Russian and USA age-structured survey results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8
# Display plots
p5 / p6
```

```{r}
#| eval: true
#| label: fig-plots_comparison3
#| fig-cap: "Various comparisons between Russian and USA age-structured survey results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8
# 4. DETAILED STATISTICAL SUMMARY
cat("\n=== DETAILED STATISTICS ===\n")

# Summary statistics by country and age
stats_summary <- combined_data %>%
  group_by(Country, Age) %>%
  summarise(
    Mean = mean(Numbers),
    Median = median(Numbers),
    SD = sd(Numbers),
    CV = sd(Numbers) / mean(Numbers),
    Min = min(Numbers),
    Max = max(Numbers),
    .groups = "drop"
  ) %>%
  mutate(across(c(Mean, Median, SD, Min, Max), ~round(.x, 1)),
         CV = round(CV, 3))

cat("Summary statistics by country and age:\n")
stats_summary %>%
  arrange(Country, Age) %>%
  print(n = 20)

# Coefficient of variation comparison
cv_comparison <- stats_summary %>%
  select(Country, Age, CV) %>%
  pivot_wider(
    names_from = Country,
    values_from = CV,
    names_glue = "{Country}_CV"
  ) %>% arrange(Age)

cat("\nCoefficient of Variation by age:\n")
print(cv_comparison)

# Correlation analysis results
high_corr_ages <- correlation_df %>% filter(Correlation > 0.5) %>% pull(Age)
low_corr_ages <- correlation_df %>% filter(Correlation < 0.2) %>% pull(Age)

cat("\nAge classes with correlation > 0.5:", paste(paste0("Age", high_corr_ages), collapse = ", "), "\n")
cat("Age classes with correlation < 0.2:", paste(paste0("Age", low_corr_ages), collapse = ", "), "\n")

combined_data <- bind_rows(russian_long, usa_long)

# Calculate correlations using wide format for easier computation
russian_wide <- combined_data %>% 
  filter(Country == "Russia") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

usa_wide <- combined_data %>% 
  filter(Country == "USA") %>% 
  select(-Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

# Age-specific correlations
age_correlations <- map_dbl(3:12, ~cor(russian_wide[[.x]], usa_wide[[.x]]))
names(age_correlations) <- paste0("Age", 1:10)

# cat("Correlations by Age Class:\n")
 print(round(age_correlations, 3))

# Overall correlation
all_russian <- russian_wide %>% select(-Source,-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
all_usa <- usa_wide %>% select(-Source,-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
overall_correlation <- cor(all_russian, all_usa)
overall_corr <- paste0("Overall correlation across all ages:", round(overall_correlation, 3), "")
# Create correlation matrix for ggplot correlation plot
correlation_matrix_data <- combined_data %>% select(Year, Age, Numbers, Country) |> 
  pivot_wider(names_from = Country, values_from = Numbers) %>%
  select(-Year) %>%
  group_by(Age) %>%
  summarise(
    Correlation = cor(Russia, USA, use = "complete.obs"),
    .groups = "drop"
  )

```

## Results comparing Russian and USA age-structured assessments

`r print(overall_corr)`.

## Comparison with the Gulf of Alaska (GOA) pollock estimates {#sec-ebsgoa}

Since the Russian context suggest some connectivity, for perspective, we thought it might
be useful to conduct a similar comparison with results from the GOA pollock stock. Pollock
assessments from the eastern Bering Sea (EBS) and Gulf of Alaska (GOA) show strong
contrasts in both magnitude and dynamics of numbers-at-age. The EBS consistently supports
far higher abundances across all ages and years (typically 5–15 times greater than the
GOA; @fig-plots_comparison_goa). Recruitment pulses are evident in both regions,
particularly at age 1, but the correlations by age class reveal only modest positive
associations at younger ages, while older ages show little to no correspondence
(@tbl-goa_corr). This indicates that early life stages may share some environmental
drivers across basins, but overall population dynamics remain largely decoupled.

A more detailed breakdown by age class underscores these findings
(@fig-plots_comparison_goa2). Cohort strength in the EBS exhibits large variability with
recurring strong year classes, while the GOA maintains lower overall abundance with weaker
pulses. Time-series plots for ages 1–6 highlight this contrast, with the EBS showing sharp
recruitment spikes and the GOA remaining more muted.

```{r}
#| eval: true
#| label: tbl-goa_corr
#| tbl-cap: "Summary of age-specific correlations between GOA and EBS age classes, 1995-2024."
#| echo: false
#| message: false
#| warning: false

df_goa <- read_csv(here::here("doc","data","goapoll.csv")) |> filter(Year>=1995)

goa_data <- df_goa %>% mutate(Source = "GOA") %>% select(Year, starts_with("Age"))
# Create tidy data format
goa_long <- df_goa |> 
  pivot_longer(cols = starts_with("Age"), names_to = "Age", values_to = "Numbers") %>%
  mutate(Region = "GOA",
         Age = str_extract(Age, "\\d+") %>% as.numeric())

combined_data <- bind_rows(goa_long |> mutate(Region="GOA"), usa_long |> mutate(Region="EBS"))
correlation_matrix_data <- combined_data %>% select(Year, Age, Numbers, Region) |> 
  pivot_wider(names_from = Region, values_from = Numbers) %>%
  select(-Year) %>%
  group_by(Age) %>%
  summarise(
    Correlation = cor(GOA, EBS, use = "complete.obs"),
    .groups = "drop"
  )
totals_data <- combined_data %>% group_by(Year, Region) %>% summarise(Total = sum(Numbers, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Region, values_from = Total) %>% mutate(Ratio_EBS_to_GOA = EBS / GOA)

goa_wide <- combined_data %>%  filter(Region == "GOA") %>%  select(-Region,-Source, -Country) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

ebs_wide <- combined_data %>%  filter(Region == "EBS") %>%  select(-Country, -Source,-Region) %>% 
  pivot_wider(names_from = Age, values_from = Numbers, names_prefix = "Age")

# Age-specific correlations
age_correlations <- map_dbl(2:11, ~cor(goa_wide[[.x]], ebs_wide[[.x]]))
names(age_correlations) <- paste0("Age", 1:10)

#cat("Correlations by Age Class:\n")
#print(round(age_correlations, 3)) 
# Overall correlation
all_goa <- goa_wide %>% select(-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
all_ebs <- ebs_wide %>% select(-Year) %>% as.matrix() %>% as.vector() |> as.numeric()
overall_correlation <- cor(all_goa, all_ebs, use = "complete.obs")
#cat("\nOverall correlation across all ages:", round(overall_correlation, 3), "\n")

# Create correlation dataframe for plotting
correlation_df <- tibble( Age = 1:10, Correlation = age_correlations )

gt(correlation_df) %>% tab_header( title = "Correlations between GOA and EBS Age Classes" ) %>%
  fmt_number( columns = vars(Correlation), decimals = 3 )


```

```{r}
#| eval: true
#| label: fig-plots_comparison_goa
#| fig-cap: "Various comparisons between GOA and EBS age-structured assessment results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8

# Plot 1: Total numbers over time
p1 <- totals_data %>%
  select(Year, GOA, EBS) %>%
  pivot_longer(cols = c(GOA, EBS), names_to = "Region", values_to = "Total") %>%
  ggplot(aes(x = Year, y = Total, color = Region)) +
  geom_line(size = 1.2) + geom_point(alpha = 0.7) +
  scale_y_continuous(labels = comma_format(), limits = c(0,NA)) +
  scale_color_manual(values = c("GOA" = "#E31A1C", "EBS" = "#1F78B4")) +
  labs(title = "Total Numbers at Age Over Time", subtitle = "Comparison between GOA and EBS assessments",
       x = "Year", y = "Total Numbers (thousands)", color = "Region") +
  theme_minimal() + ylim(c(0,NA)) + theme(plot.title = element_text(size = 14, face = "bold"), legend.position = "top")

# Plot 2: Age 1 comparison over time
p2 <- combined_data %>% filter(Age == 1) %>%
  ggplot(aes(x = Year, y = Numbers, color = Region)) +
  geom_line(size = 1.2) + geom_point(alpha = 0.7) +
  scale_y_continuous(labels = comma_format(), limits=c(0,NA)) +
  scale_color_manual(values = c("GOA" = "#E31A1C", "EBS" = "#1F78B4")) +
  labs(title = "Age 1 Numbers Over Time", x = "Year", y = "Numbers (thousands)",
       color = "Region") + theme_minimal() + theme(legend.position = "top")

# Plot 3: EBS/GOA ratio over time
p3 <- totals_data %>%
  ggplot(aes(x = Year, y = Ratio_EBS_to_GOA)) +
  geom_line(size = 1.2, color = "#FF7F00") +
  geom_point(alpha = 0.7, color = "#FF7F00") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50", alpha = 0.7) +
  labs(title = "EBS/GOA Total Ratio Over Time",
       x = "Year", y = "Ratio (EBS/GOA)") +ylim(c(0,NA)) +
  theme_minimal()

# Plot 4: Correlations by age class
p4 <- correlation_df %>%
  ggplot(aes(x = factor(Age), y = Correlation)) + geom_col(fill = "#33A02C", alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = c(0.5, -0.5), linetype = "dotted", alpha = 0.5, color = "red") +
  labs(title = "Correlations by Age Class",
       x = "Age Class", y = "Correlation Coefficient") +
  ylim(-1, 1) +
  theme_minimal()

# Combine plots using patchwork
(p1 | p2) / (p3 | p4)

```

```{r}
#| eval: true
#| label: fig-plots_comparison_goa2
#| fig-cap: "Various comparisons between GOA and EBS age-structured assessment results."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8

# Plot 5: Age composition comparison (heatmap style)
p5 <- combined_data |>  group_by(Region, Age) |>  summarise(Mean_Numbers = mean(Numbers), .groups = "drop") %>%
  ggplot(aes(x = factor(Age), y = Region, fill = Mean_Numbers)) + geom_tile() +
  scale_fill_gradient2(low = "white", high = "#2166AC",  trans = "log10", labels = comma_format()) +
  labs(title = "Mean Numbers by Age Class", subtitle = "Log-scale heatmap of average numbers across all years", x = "Age Class", y = "Region", fill = "Mean\nNumbers") +
  theme_minimal()

# Plot 6: Time series by age class (faceted)
p6 <- combined_data %>% filter(Age <= 6) %>%  
  ggplot(aes(x = Year, y = Numbers, color = Region)) +
  geom_line(size = 1) + facet_wrap(~paste("Age", Age), scales = "free_y") +
  scale_y_continuous(labels = comma_format()) + scale_color_manual(values = c("GOA" = "#E31A1C", "EBS" = "#1F78B4")) +
  labs(title = "Numbers at Age Over Time (Ages 1-6)", x = "Year", y = "Numbers (thousands)", color = "Region") +
  theme_minimal() + theme(legend.position = "top", strip.text = element_text(size = 10, face = "bold"))

p5 / p6

```

# Other items

The CIE noted in particular that some documentation of aspects of the ADMB code was
incomplete. During the meetings, we began to more properly specify how the selectivity
penalties were being calculated and implementated.

### First‑difference selectivity penalty (temporal) {#test-fd-penalty-time}

We need the penalty used in ADMB where the **first year is compared to zero** (not to a
previous estimated year). For a matrix of log‑selectivities ($\log s_{y,a}$)
(years$(y=1,\ldots,T)$, ages $(a=1,\ldots,A)$, the penalty is:

$$
\mathcal{P}_{\text{FD,time}} \,=\, w \sum_{a=1}^{A}
\Big[
\big(\log s_{1,a}-0\big)^2
+ \sum_{y=2}^{T}\big(\log s_{y,a}-\log s_{y-1,a}\big)^2
\Big].
$$

#### RTMB/R implementation

```{r}
#| echo: true
# log_sel: matrix [T x A] of log-selectivities (year rows, age cols)
# w: scalar or length-A vector of weights
fd_penalty_time <- function(log_sel, w = 1) {
  stopifnot(is.matrix(log_sel))
  T <- nrow(log_sel); A <- ncol(log_sel)
  if (length(w) == 1) w <- rep(w, A)
  stopifnot(length(w) == A)

  diffs <- rbind(log_sel[1, , drop = FALSE] - 0,   # year 1 vs 0
                 apply(log_sel, 2, diff))          # y>=2 diffs
  sum(colSums((sqrt(w) * diffs)^2))
}
```

### Sanity check

```{r}
#| echo: true
set.seed(42)
T <- 5; A <- 3
L <- matrix(rnorm(T*A, sd=0.2), T, A); w <- c(1, 2, 0.5)

fd_loop <- function(L, w){
  acc <- 0
  for (a in seq_len(ncol(L))) {
    acc <- acc + w[a]*(L[1,a]-0)^2
    for (y in 2:nrow(L)) acc <- acc + w[a]*(L[y,a]-L[y-1,a])^2
  }
  acc
}

all.equal(fd_penalty_time(L, w), fd_loop(L, w))
```

### First‑difference selectivity penalty (age) {#test-fd-penalty-age}

For within-year smoothing over age with a **zero anchor at the minimum age**:

$$
\mathcal{P}_{\text{FD,age}} \,=\, w \sum_{y=1}^{T}
\Big[
\big(\log s_{y,a_{\min}}-0\big)^2
+ \sum_{a=a_{\min}+1}^{A}\big(\log s_{y,a}-\log s_{y,a-1}\big)^2
\Big].
$$

#### RTMB/R implementation

```{r}
#| echo: true
# log_sel: matrix [T x A] of log-selectivities (year rows, age cols)
# a_min: column index of minimum age within 'log_sel' (default = 1)
fd_penalty_age <- function(log_sel, w = 1, a_min = 1) {
  stopifnot(is.matrix(log_sel))
  T <- nrow(log_sel); A <- ncol(log_sel)
  stopifnot(a_min >= 1, a_min <= A)
  if (length(w) == 1) w <- rep(w, T)
  stopifnot(length(w) == T)

  # Build diffs across age with anchor at a_min (assumes a_min=1; if not, set zeros below a_min)
  diffs <- matrix(0, nrow = T, ncol = A)
  diffs[, a_min] <- log_sel[, a_min] - 0
  if (a_min < A) {
    for (a in (a_min+1):A) {
      diffs[, a] <- log_sel[, a] - log_sel[, a-1]
    }
  }
  sum(rowSums((sqrt(w) * diffs)^2))
}
```

## Reference: survey lognormal scaling (consistency check) {#test-survey-lognorm}

When porting lognormal survey likelihoods, confirm whether ADMB’s objective includes the
normalizing constant and whether (\sigma) is on the **log** scale:

$\ell(\sigma) = -\sum_i \frac{(\log O_i - \log E_i)^2}{2\sigma^2} \; -\; n\log\sigma \; -\; \frac{n}{2}\log(2\pi).$

```{r}
#| echo: true
lognorm_nll <- function(obs, exp, sigma_log, include_const = TRUE) {
  r <- log(pmax(obs, .Machine$double.eps)) - log(pmax(exp, .Machine$double.eps))
  n <- length(r)
  nll <- sum(r^2) / (2 * sigma_log^2)
  if (include_const) nll <- nll + n*log(sigma_log) + n*0.5*log(2*pi)
  nll
}
```

# References
